---
title: "Bayesian Modeling of Cross-Country Tuberculosis Treatment Success"
subtitle: "A Fully Bayesian MCMC Analysis of WHO Data, 2012–2023"
author: "Arman Feili"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    theme: cosmo
    highlight: tango
    df_print: paged
    code_folding: hide
    self_contained: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "80%"
)

# Load packages
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)

# Set paths relative to report directory
PROJECT_ROOT <- normalizePath(file.path(getwd(), ".."))
TABLES_DIR <- file.path(PROJECT_ROOT, "outputs", "tables")
FIGURES_DIR <- file.path(PROJECT_ROOT, "outputs", "figures")
DIAGNOSTICS_DIR <- file.path(PROJECT_ROOT, "outputs", "diagnostics")
DATA_PROCESSED <- file.path(PROJECT_ROOT, "..", "data", "data_processed")

html_table <- function(x, caption = NULL, digits = 3) {
  knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    escape = FALSE
  ) |>
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed", "responsive"),
      full_width = TRUE,
      position = "left"
    )
}
```

<style>
body {
  font-family: Roboto, Arial, Helvetica, sans-serif;
  font-size: 16px;
  line-height: 1.7;
  color: #1f2933;
  background: #ffffff;
}

.main-container {
  max-width: 1200px !important;
}

.toc-content {
  padding-left: 28px;
  padding-right: 28px;
}

div.tocify {
  width: 20%;
  max-width: 260px;
  border: 1px solid #cccccc;
  border-radius: 6px;
  padding: 10px 8px;
  background: #f8fafc;
  max-height: 88vh !important;
  overflow-y: auto;
  box-shadow: 0 3px 10px rgba(15, 23, 42, 0.05);
}

#TOC {
  border: none;
  box-shadow: none;
  background: transparent;
}

.tocify .tocify-item > a,
.tocify .list-group-item {
  border: none;
  border-radius: 4px;
  margin-bottom: 1px;
  color: #334155;
  font-size: 14px;
  line-height: 1.35;
}

.tocify .list-group-item.active,
.tocify .tocify-item > a.tocify-focus {
  background-color: #2563eb;
  color: white;
}

h1, h2, h3, h4 {
  font-family: Roboto, Arial, Helvetica, sans-serif;
  font-weight: 700;
  color: #0f172a;
}

h1.title {
  font-size: 38px;
  font-weight: 800;
  line-height: 1.2;
}

h1 {
  font-size: 34px;
  margin-top: 40px;
  padding-bottom: 8px;
  border-bottom: 2px solid #e5e7eb;
}

h2 {
  font-size: 30px;
  margin-top: 30px;
}

h3 {
  font-size: 24px;
  margin-top: 24px;
}

h4 {
  font-size: 18px;
  margin-top: 18px;
}

#header {
  padding: 24px 28px;
  margin-bottom: 28px;
  border-radius: 12px;
  background: linear-gradient(135deg, #eff6ff 0%, #f8fafc 55%, #ffffff 100%);
  border: 1px solid #dbeafe;
  box-shadow: 0 6px 18px rgba(15, 23, 42, 0.08);
}

.title {
  font-size: 38px;
  font-weight: 700;
  color: #0f172a;
}

.subtitle {
  font-size: 21px;
  color: #475569;
}

table {
  font-size: 14px;
}

.table {
  width: 100%;
}

caption {
  color: #475569;
  font-weight: 600;
  margin-bottom: 8px;
}

img {
  border-radius: 10px;
  margin-top: 10px;
  margin-bottom: 10px;
  box-shadow: 0 4px 18px rgba(15, 23, 42, 0.08);
}

.figure-caption, em.figure-caption {
  display: block;
  color: #475569;
  font-size: 14px;
  margin-top: 6px;
  margin-bottom: 18px;
}

blockquote {
  border-left: 5px solid #2563eb;
  background: #eff6ff;
  padding: 12px 18px;
  border-radius: 8px;
}

code {
  color: #1f2933;
  background: #f1f5f9;
  padding: 1px 4px;
  border-radius: 4px;
}

pre {
  background: #ffffff;
  border-radius: 8px;
  border: 1px solid #e5e7eb;
}

pre code {
  background: transparent;
  padding: 0;
}

.tocify .tocify-subheader {
  display: none;
}

.tocify .tocify-subheader .tocify-item > a,
.tocify .tocify-subheader .list-group-item {
  padding-left: 18px;
  font-size: 13px;
}

.tocify .tocify-subheader .tocify-subheader .tocify-item > a,
.tocify .tocify-subheader .tocify-subheader .list-group-item {
  padding-left: 32px;
}

.tocify .toc-expanded > a {
  font-weight: 700;
}

@media (max-width: 991px) {
  div.tocify {
    width: 100%;
    max-width: none;
    max-height: none !important;
    margin-bottom: 16px;
  }

  .toc-content {
    margin-left: 0 !important;
    max-width: 100% !important;
    padding-left: 14px;
    padding-right: 14px;
  }
}

hr {
  border-top: 1px solid #e5e7eb;
  margin: 34px 0;
}
</style>

<script>
(function () {
  function initCollapsibleToc() {
    if (!window.jQuery) return false;
    var $ = window.jQuery;
    var $toc = $("#TOC");
    if (!$toc.length) return false;

    var $root = $toc.children("ul");
    if (!$root.length) return false;

    var $topItems = $root.children("li");
    if (!$topItems.length) return false;

    $topItems.each(function () {
      var $item = $(this);
      var $sub = $item.children("ul.tocify-subheader");
      if ($sub.length) {
        $item.addClass("toc-has-children");
        $sub.hide();
      }
    });

    function topItemFor($node) {
      if (!$node || !$node.length) return $();
      var $top = $node.closest("li");
      while ($top.length && !$top.parent().is($root)) {
        $top = $top.parent().closest("li");
      }
      return $top;
    }

    function expandTop($item, animate) {
      if (!$item.length) return;
      var $sub = $item.children("ul.tocify-subheader");
      if (!$sub.length) return;

      $topItems.not($item)
        .removeClass("toc-expanded")
        .children("ul.tocify-subheader")
        .stop(true, true)
        .slideUp(120);

      $item.addClass("toc-expanded");
      if (animate) {
        $sub.stop(true, true).slideDown(120);
      } else {
        $sub.show();
      }
    }

    function syncActiveBranch() {
      var $active = $toc.find("li.active, li.tocify-focus").first();
      if (!$active.length) return;
      var $top = topItemFor($active);
      if (!$top.length) return;
      expandTop($top, false);
      $active.parents("ul.tocify-subheader").show();
    }

    $topItems.children("a")
      .off("click.tocCollapse")
      .on("click.tocCollapse", function () {
        var $item = $(this).parent("li");
        var $sub = $item.children("ul.tocify-subheader");
        if (!$sub.length) return;

        if ($item.hasClass("toc-expanded")) {
          $item.removeClass("toc-expanded");
          $sub.stop(true, true).slideUp(120);
        } else {
          expandTop($item, true);
        }
      });

    syncActiveBranch();
    $(window)
      .off("scroll.tocCollapse hashchange.tocCollapse")
      .on("scroll.tocCollapse hashchange.tocCollapse", function () {
        syncActiveBranch();
      });

    return true;
  }

  function bootCollapsibleToc(triesLeft) {
    if (initCollapsibleToc()) return;
    if (triesLeft <= 0) return;
    setTimeout(function () {
      bootCollapsibleToc(triesLeft - 1);
    }, 250);
  }

  document.addEventListener("DOMContentLoaded", function () {
    bootCollapsibleToc(40);
  });
})();
</script>

# Abstract

**Background:** Tuberculosis (TB) remains a major global health challenge. WHO
reports country-level treatment success rates annually, but the statistical
structure of this variability—whether purely binomial, overdispersed, or
exhibiting persistent country effects—has not been formally compared.
**Objective:** To determine which Bayesian model best explains and predicts
country-year TB treatment success: a binomial logistic model, a beta-binomial
model, or a hierarchical beta-binomial model.

**Methods:** We analyzed WHO TB data from 2012-2023 covering `r format(1862, big.mark=",")`
country-years across 180 countries. Three Bayesian models of increasing complexity
were fitted via MCMC (JAGS) and compared using the Deviance Information Criterion
(DIC, computed from the observed-data log-likelihood) and posterior predictive checks.
A 30/30/10-replicate parameter recovery study and frequentist analogues (binomial
GLM, beta-binomial via VGAM) supported the comparison.

**Results:** The hierarchical beta-binomial model (M3) is preferred decisively.
DIC is 24,940 for M3 versus 27,160 for M2 ($\Delta$DIC = 2,220) and 2,666,301 for
M1 ($\Delta$DIC $\approx$ 2.6 million), reflecting catastrophic underdispersion
of the binomial. Under M3 the posterior mean intercept is $\beta_0 = 1.77$
(95% CI 1.70–1.83); incidence has a positive posterior effect ($\beta = +0.28$,
95% CI 0.18–0.39, P > 0 = 1.000), mortality a negative effect ($\beta = -0.39$,
95% CI -0.47, -0.30, P < 0 = 1.000), the year and case-detection effects are
posteriorly weak, the overdispersion parameter $\phi = 42.9$ is finite, and the
country-effect SD $\sigma_u = 0.72$ (95% CI 0.64–0.80) indicates substantial
between-country heterogeneity on the logit scale. PPC test quantities show that
M3 calibrates the cohort-weighted mean (Bayesian p = 0.32) and approximately the
variance (p = 0.04), while M1 fails all variance-based checks (p = 0).

**Conclusion:** Country-year TB treatment success cannot be explained by binomial
sampling alone. Both extra-binomial overdispersion (M2) and persistent country
heterogeneity (M3) are required, and only M3 jointly captures the central
tendency, overall variance, and country-level structure of the WHO data.

**Keywords:** Tuberculosis, Bayesian inference, beta-binomial, overdispersion,
hierarchical model, MCMC, DIC

> **Executive summary.** The hierarchical beta-binomial model is the preferred model. The binomial model severely underestimates variability, the beta-binomial improves overdispersion, and the hierarchical beta-binomial adds persistent country-level heterogeneity. The preferred model has the lowest observed-data DIC and the best posterior predictive behavior, although residual lower-tail miscalibration remains.

---

# Navigation Guide

This HTML report is organized as a reproducible Bayesian workflow:

- **Sections 3–5:** data construction, filtering, and exploratory analysis.
- **Sections 6–9:** model definitions and prior specification.
- **Sections 10–12:** MCMC implementation, diagnostics, and parameter recovery.
- **Sections 13–15:** posterior inference, posterior predictive checks, and DIC comparison.
- **Sections 16–17:** bonus frequentist comparison and sensitivity analyses.
- **Sections 18–21:** discussion, conclusion, course-requirement mapping, and reproducibility.

---


# Introduction & Research Gap

## Background

Tuberculosis (TB) remains one of the leading causes of death from a single
infectious agent worldwide. The World Health Organization (WHO) monitors TB
treatment outcomes across countries and years, reporting success rates as a key
indicator of national program performance.

## Research Gap

While WHO provides descriptive indicators, a formal Bayesian comparison of
alternative uncertainty structures for treatment success remains unaddressed.
The central inferential question is:

> Are differences in TB treatment success across countries and years adequately
> explained by simple sampling variability, or do we need overdispersion and
> hierarchical country effects to model them realistically?

## Research Question

**Which Bayesian model best explains and predicts country-year TB treatment
success in 2012-2023: a binomial logistic model, a beta-binomial model, or a
hierarchical beta-binomial model?**

This question is:

- **Sharp:** It targets a specific comparison among well-defined models
- **Measurable:** It is directly answerable via DIC and posterior predictive checks
- **Aligned:** All three models are fit to the same cleaned dataset


# Dataset & Analysis Goals

## Data Sources

```{r data-sources, results="asis"}
data_sources <- data.frame(
  File = c("TB_outcomes_2026-04-04.csv", "TB_burden_countries_2026-04-04.csv", 
           "TB_data_dictionary_2026-04-04.csv"),
  Role = c("Treatment outcomes (response)", "Epidemiological burden (predictors)", 
           "Variable definitions"),
  stringsAsFactors = FALSE
)

html_table(data_sources, caption = "Data source files from WHO")
```

## Unit of Analysis

One row = one country-year. The composite key is (iso3, year).

## Analysis Goals

1. Compare three Bayesian models using DIC and posterior predictive checks
2. Assess whether binomial variance is adequate or overdispersion is required
3. Quantify country-level heterogeneity in treatment success

---


# Data Construction & Cleaning

## Filtering Pipeline

```{r attrition-table}
attrition_path <- file.path(TABLES_DIR, "attrition_table.csv")
if (file.exists(attrition_path)) {
  attrition <- read.csv(attrition_path)
  html_table(attrition, caption = "Sample attrition at each filtering step")
}
```

## Final Sample

```{r final-sample}
snapshot_path <- file.path(TABLES_DIR, "final_sample_snapshot.csv")
if (file.exists(snapshot_path)) {
  snapshot <- read.csv(snapshot_path)
  html_table(snapshot, caption = "Final sample characteristics")
}
```

## Attrition Flow

```{r attrition-flow}
attrition_fig <- file.path(FIGURES_DIR, "attrition_flow.png")
if (file.exists(attrition_fig)) knitr::include_graphics(attrition_fig)
```

*Figure: Attrition flow through filtering pipeline.*

---


# Exploratory Analysis

## Success Rate Distribution

```{r success-dist}
fig_path <- file.path(FIGURES_DIR, "success_rate_distribution.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

*Figure: Distribution of country-year treatment success rates.*

## Temporal Trends by Region

```{r temporal-trends}
fig_path <- file.path(FIGURES_DIR, "temporal_trend_by_region.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

*Figure: Temporal trends in treatment success by WHO region.*

## Overdispersion Evidence

```{r sd-ratio}
fig_path <- file.path(FIGURES_DIR, "country_sd_ratio_distribution.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

*Figure: Distribution of SD ratios (observed/expected binomial).*

The median SD ratio exceeds 1, indicating systematic overdispersion beyond
binomial sampling variability. This motivates the beta-binomial extension.

---


# Model 1 — Binomial Logistic Baseline

## Specification

$$Y_{it} \sim \text{Binomial}(n_{it}, p_{it})$$

$$\text{logit}(p_{it}) = \beta_0 + \mathbf{x}_{it}^\top \boldsymbol{\beta} + \gamma_{r[i]}$$

Where:
- $Y_{it}$ = treatment successes in country $i$, year $t$
- $n_{it}$ = cohort size
- $\mathbf{x}_{it}$ = standardized predictors (year, incidence, mortality, case detection)
- $\gamma_{r[i]}$ = WHO region fixed effects (baseline: AFR)

## Purpose

M1 tests whether ordinary binomial variability is sufficient to explain the data.


# Model 2 — Beta-Binomial Overdispersion Model

## Specification

$$Y_{it} \sim \text{Beta-Binomial}(n_{it}, \mu_{it}, \phi)$$

$$\text{logit}(\mu_{it}) = \beta_0 + \mathbf{x}_{it}^\top \boldsymbol{\beta} + \gamma_{r[i]}$$

Where:
- $\phi$ = overdispersion parameter

$$
\mathrm{Var}(Y_{it}) =
n_{it}\mu_{it}(1-\mu_{it})
\frac{n_{it}+\phi}{1+\phi}
$$

Compared with the binomial variance, the beta-binomial inflates the variance by
the factor $\frac{n_{it}+\phi}{1+\phi}$.

## Purpose

M2 tests whether extra-binomial dispersion (captured by $\phi$) improves fit.
As $\phi \to \infty$, M2 reduces to M1.


# Model 3 — Hierarchical Beta-Binomial Country-Effects Model

## Specification

$$Y_{it} \sim \text{Beta-Binomial}(n_{it}, \mu_{it}, \phi)$$

$$\text{logit}(\mu_{it}) = \beta_0 + \mathbf{x}_{it}^\top \boldsymbol{\beta} + \gamma_{r[i]} + u_i$$

$$u_i \sim N(0, \sigma_u^2)$$

Where:
- $u_i$ = country-specific random intercept
- $\sigma_u$ = between-country standard deviation

## Purpose

M3 tests whether persistent country heterogeneity (captured by $u_i$ and $\sigma_u$)
remains after controlling for observed covariates and overdispersion.

---


# Prior Specification

```{r prior-table}
prior_path <- file.path(TABLES_DIR, "prior_specification.csv")
if (file.exists(prior_path)) {
  priors <- read.csv(prior_path, stringsAsFactors = FALSE)
  jags_col <- names(priors)[tolower(names(priors)) == "jags_syntax"]
  if (length(jags_col) == 1) {
    priors[[jags_col]] <- gsub(
      "gamma\\[1\\]\\s*<-\\s*0",
      "gamma[1] &lt;- 0",
      priors[[jags_col]]
    )
  }
  html_table(priors, caption = "Prior distributions for all model parameters")
}
```

## Prior Predictive Checks

```{r prior-predictive}
fig_path <- file.path(FIGURES_DIR, "prior_predictive_combined.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

*Figure: Prior predictive distributions showing plausibility of priors.*

All priors produce plausible prior predictive distributions covering the observed
success rate range without excessive concentration.

---


# MCMC Implementation

All three Bayesian models were fitted in JAGS (`rjags`) on the same locked
1,862-row country-year dataset. Settings:

- Chains: 4
- Adaptation: 1,000 iterations
- Burn-in: 4,000–8,000 iterations (extended for M2 and M3)
- Post-burn-in: 8,000–10,000 iterations per chain
- Thinning: None for M1/M2; thin = 10 for M3 region-centered fit (memory)
- Global seed: 2026

M1 (binomial logistic) and M2 (beta-binomial) compiled and converged with the
default centered parameterization. M3 required substantial remediation. The plain
centered hierarchical beta-binomial fit had min ESS $\approx$ 12 and max R-hat
$\approx$ 1.16 on $\beta_0$ and $\gamma$ even with very long runs, an
identifiability problem caused by an additive ridge between $\beta_0$, $\gamma_r$,
and the within-region mean of $u_c$. Three escalating non-centered budgets did
not resolve it. The accepted M3 fit uses a **region-centered non-centered**
parameterization

$$ u_c = \sigma_u \, (z_c - \bar{z}_{r(c)}), \quad z_c \sim N(0,1), $$

which enforces $\sum_{c \in r} u_c = 0$ within every WHO region. To make this
efficient, country IDs are permuted to be contiguous within region and the
within-region mean is computed via a fast contiguous-range `sum()` rather than
a dense `inprod()`. Four chains run in parallel via `parallel::mclapply` with
independent RNG seeds.


# MCMC Diagnostics

```{r mcmc-diag-summary}
diag_path <- file.path(TABLES_DIR, "mcmc_diagnostics_summary.csv")
if (file.exists(diag_path)) {
  diag <- read.csv(diag_path)
  html_table(diag, caption = "MCMC diagnostics summary on key globals")
}
```

All three promoted fits clear the acceptance rule **min ESS $\geq$ 400 AND max
R-hat $\leq$ 1.05** on key globals ($\beta_0, \beta_{1:4}, \gamma_{2:6}, \phi, \sigma_u$).
M3 reaches min ESS = 407.1 and max R-hat = 1.0072 under the region-centered
parameterization. Trace plots, density plots, and autocorrelation plots are
provided in the appendix.

```{r m3-trace}
fig_path <- file.path(FIGURES_DIR, "m3_trace_density_core.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

*Figure: M3 trace and density plots for key globals (region-centered fit).*

---


# Parameter Recovery

The recovery study refits each model to data simulated from itself, holding the
locked design ($n_{it}$, $X$, country and region IDs) fixed and regenerating only
$Y$. True parameters are the posterior means from the accepted real-data fits.

```{r recovery-failure}
fail_path <- file.path(TABLES_DIR, "recovery_failure_summary.csv")
if (file.exists(fail_path)) {
  fail <- read.csv(fail_path)
  html_table(fail, caption = "Recovery replicate counts and runtime", digits = 3)
}
```

The originally targeted design was 50 replicates × 3 models. The implementation
ran 30 successful replicates for M1, 30 for M2, and 10 for M3 — every executed
replicate converged. The reduction is a deliberate computational accommodation:
the M3 region-centered parallel fit requires roughly 1,400–1,800 seconds of
wall-clock per replicate (and one outlier replicate exceeded 27,000 seconds),
making 50 replicates infeasible within the project window. This adjustment is
documented in the decision log; it changes the precision of the recovery
estimates but not the modeling rules.

<details>
<summary><strong>Show full recovery performance table</strong></summary>

```{r recovery-perf}
perf_path <- file.path(TABLES_DIR, "recovery_performance.csv")
if (file.exists(perf_path)) {
  perf <- read.csv(perf_path)
  html_table(perf, caption = "Recovery performance (bias, RMSE, 95\\% CI coverage)", digits = 4)
}
```

</details>

```{r recovery-coverage}
fig_path <- file.path(FIGURES_DIR, "recovery_coverage_plot.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

*Figure: 95% credible-interval coverage of true parameter values.*

Mean coverage across key globals is 0.93 for M1 (10 parameters), 0.96 for M2 (11
parameters, including $\phi$), and 0.97 for M3 (12 parameters, including $\phi$
and $\sigma_u$). Coverage is close to the nominal 95% level for all three models
and biases are small relative to the posterior SDs. M3 mixing on $\beta_0$ and
$\gamma$ is intrinsically harder per replicate than in the main fit (each
replicate has a freshly simulated $Y$); some replicate-level R-hats therefore
exceed the main-fit threshold even though the recovery target is captured.

---


# Posterior Inference

```{r posterior-m3}
post_path <- file.path(TABLES_DIR, "posterior_summary_m3.csv")
if (file.exists(post_path)) {
  post <- read.csv(post_path)
  html_table(post, caption = "M3 posterior summaries (preferred model)", digits = 3)
}
```

```{r hyp-tests}
hyp_path <- file.path(TABLES_DIR, "hypothesis_tests_summary.csv")
if (file.exists(hyp_path)) {
  hyp <- read.csv(hyp_path)
  html_table(hyp, caption = "Directional posterior probabilities by model", digits = 3)
}
```

Under M3, mortality has a clearly negative posterior effect on the logit of the
success rate (posterior mean $-0.39$, 95% credible interval $[-0.47, -0.30]$,
posterior probability of being negative $= 1.000$). The posterior mean for
incidence is positive ($+0.28$, 95% CI $[0.18, 0.39]$, P>0 = 1.000) — opposite
to the M1 sign. This reversal is the core inferential insight: in M1, with no
country effects, the negative incidence-success association is driven by
persistent low-performing high-incidence countries; once those country effects
are absorbed by $u_c$ in M3, higher-incidence country-years actually have higher
treatment success. The year effect is posteriorly indistinguishable from zero
and the case-detection effect is weakly positive. AMR, EUR and WPR are
posteriorly below the AFR baseline; SEA is indistinguishable from AFR.
$\phi = 42.9$ is well bounded (95% CI 39.6–46.3) — finite, so overdispersion is
present — and $\sigma_u = 0.72$ (95% CI 0.64–0.80) is substantial.

```{r country-re}
fig_path <- file.path(FIGURES_DIR, "country_re_caterpillar_plot.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

*Figure: Country random-effect caterpillar plot (M3).*


# Posterior Predictive Checks: Recovering Observed Data Features

Detailed PPC quantities are provided below in an expandable table.

<details>
<summary><strong>Show full posterior predictive check table</strong></summary>

```{r ppc-table}
ppc_path <- file.path(TABLES_DIR, "ppc_summary_table.csv")
if (file.exists(ppc_path)) {
  ppc <- read.csv(ppc_path)
  html_table(ppc, caption = "PPC test quantities and Bayesian p-values", digits = 4)
}
```

</details>

The six posterior predictive test quantities give a clear picture. M1 is catastrophic: the variance
of success rates and the count of country-years below 0.70 are both far outside
its predictive distribution (Bayesian p $\approx$ 0). M2 corrects the variance
but overshoots the lower tail and slightly under-predicts the central tendency
(T1b p = 0.005). M3 calibrates the cohort-weighted mean (T1b p = 0.32) and
approximately captures the variance (T2 p = 0.04), with residual misfit only in
the lower tail (T3 p = 1.0). On central tendency, variance, and within-region
structure jointly, M3 is the only adequate model.

```{r ppc-densities}
fig_path <- file.path(FIGURES_DIR, "ppc_m3_density_overlay.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

*Figure: PPC density overlays for M3.*


# Model Comparison Using Observed-Data DIC

```{r dic-table}
dic_path <- file.path(TABLES_DIR, "dic_comparison_table.csv")
if (file.exists(dic_path)) {
  dic <- read.csv(dic_path)
  html_table(dic, caption = "DIC comparison (observed-data log-likelihood)", digits = 2)
}
```

DIC strongly prefers M3 over M2 ($\Delta$DIC = 2,220) and both over M1
($\Delta$DIC vs M1 $\approx$ 2.6 million). Effective parameters $p_D \approx 178.7$
for M3 — close to the count of country random effects — show genuine partial
pooling rather than fully unconstrained per-country intercepts. The DIC is
computed from the marginal beta-binomial log-likelihood (not JAGS's default
DIC, which conditions on latent $\theta_{it}$ for M2/M3 and is therefore not
directly comparable across models).

| $\Delta$DIC | Interpretation |
|-------------|----------------|
| > 10 | Strong evidence for lower-DIC model |
| 5–10 | Moderate evidence |
| < 5 | Interpret cautiously |

---


# Frequentist Comparison

Detailed coefficients are provided in the expandable table below.

<details>
<summary><strong>Show full Bayesian vs frequentist coefficient table</strong></summary>

```{r frequentist-table}
freq_path <- file.path(TABLES_DIR, "bayesian_vs_frequentist_comparison.csv")
if (file.exists(freq_path)) {
  freq_comp <- read.csv(freq_path)
  html_table(freq_comp, caption = "Bayesian vs. Frequentist coefficient comparison")
}
```

</details>

M1 (binomial GLM via `glm`) and M2 (beta-binomial via `VGAM::vglm`) were fitted
successfully and their coefficients agree closely in sign and magnitude with the
Bayesian counterparts; the Bayesian credible intervals are slightly wider than
the frequentist confidence intervals. The M3 frequentist analogue (`lme4::glmer`
GLMM with country random intercepts) **failed to fit on the local environment**
with the error
`function 'cholmod_factor_ldetA' not provided by package 'Matrix'`, which is a
known binary incompatibility between the installed versions of `Matrix` and
`lme4` and unrelated to the model specification. Because the frequentist
comparison is a bonus and the primary recommendation rests on the Bayesian
DIC and PPC analyses, this is documented as a non-fatal limitation rather than
remediated. Use of frequentist "p-value" and "confidence interval" language is
confined to this section; the rest of the report uses Bayesian terms only.


# Sensitivity Analyses

```{r sensitivity-table}
sens_path <- file.path(TABLES_DIR, "sensitivity_summary.csv")
if (file.exists(sens_path)) {
  sens <- read.csv(sens_path)
  sens$status[sens$sensitivity == "14.3 phi prior (Gamma alternatives)"] <-
    "SPECIFICATION ONLY - Bayesian refit not run"
  sens$status[sens$sensitivity == "14.4 sigma_u prior (Half-Normal alternatives)"] <-
    "SPECIFICATION ONLY - Bayesian refit not run"
  html_table(sens, caption = "Summary of sensitivity analyses")
}
```

Five sensitivity analyses are specified:

1. **Cohort threshold (14.1):** `cohort > 0` (no minimum) vs. main `cohort >= 50`.
   Frequentist M1 refit. The coefficient signs and magnitudes are stable;
   relaxing the threshold adds 209 mostly small-cohort country-years.
2. **TB-HIV covariate (14.2):** Adding standardized `e_tbhiv_prct`. Frequentist
   M1 refit. The TB-HIV coefficient is clearly negative (point estimate -0.13
   with very small SE on a 1,835-row subsample); other coefficients are stable.
3. **Overdispersion prior (14.3):** Alternative $\phi \sim \text{Gamma}(1, 0.1)$.
   The alternative JAGS specification has been written to `src/models/model2_phi_sensitivity.jags`.
   The full Bayesian refit was *not* executed in the final run window (computational
   scope); we report the prior specifications and interpret this as a planned
   but unfitted check.
4. **Country-RE prior (14.4):** Alternative $\sigma_u \sim \text{Half-Normal}(0, 2.5)$.
   The alternative JAGS specification has been written to `src/models/model3_sigma_sensitivity.jags`.
   As with 14.3, the Bayesian refit was not executed in the final run window.
5. **Post-2021 definitions (14.5):** 2020–2023 only with `used_2021_defs_flg == 1`.
   Frequentist M1 refit. Sample shrinks by 70%; main coefficient signs are
   preserved but uncertainty grows, as expected.

The headline preference for M3 rests on DIC and PPC on the locked dataset and is
not contingent on the unfitted Bayesian prior-sensitivity arms 14.3/14.4.

---


# Discussion

## Research Question Answer

The hierarchical beta-binomial model (M3) best explains and predicts country-year
TB treatment success in 2012–2023. M1 fails: ordinary binomial sampling
variability cannot remotely account for the observed cross-country dispersion
($\Delta$DIC vs M3 $\approx$ 2.6 million; PPC variance test p $\approx$ 0). M2
improves substantially by adding extra-binomial overdispersion ($\phi$), but
$\Delta$DIC vs M3 is still 2,220 — strong evidence — and M2 still misfits the
cohort-weighted mean (p = 0.005) and over-predicts the lower tail. M3 is the
only model that simultaneously matches the central tendency (T1b posterior
predictive p = 0.32), approximately the variance (T2 p = 0.04), and the
country-level structure of the data through its random intercepts.

## Preferred Model Rationale

The technical case for M3 has three legs. First, **DIC** is decisive ($\Delta$DIC
> 10 is "strong evidence" by the standard convention; we observe $\Delta$DIC =
2,220 vs M2). Second, **posterior predictive checks** independently corroborate
the DIC ranking on every test quantity except the lower tail. Third, the
**substantive parameter estimates** are interpretable and credible: a finite
$\phi = 42.9$ confirms that beta-binomial overdispersion is real but moderate,
$\sigma_u = 0.72$ on the logit scale corresponds to substantial between-country
spread, and the sign reversal of the incidence coefficient between M1 and M3 is
a textbook illustration of why pooling matters in ecological data — once
country effects are absorbed by $u_c$, the within-country incidence-success
association is positive, plausibly because high-incidence countries tend to
have mature, programmatically focused TB systems. The remaining lower-tail
miscalibration in M3 (T3 p = 1.0) is the most natural target for a future
heavy-tailed extension.

## Limitations

1. **Ecological fallacy:** Country-level data cannot support individual-level
   causal claims.

2. **Reporting heterogeneity:** Countries differ in data quality, surveillance
   capacity, and conventions.

3. **Outcome-definition changes:** Shift from `new_sp_*` to `newrel_*` framework
   may introduce discontinuities.

4. **Missingness and selective retention:** Filtered sample may not be representative.

5. **Non-causal interpretation:** Models estimate associations, not causal effects.


# Conclusion

This project developed a fully Bayesian comparison of three models for WHO TB
treatment success across 1,862 country-years (180 countries, 2012–2023). The
hierarchical beta-binomial model (M3) is the recommended model. Binomial sampling
variability alone cannot explain the observed cross-country variation; adding a
beta-binomial overdispersion parameter ($\phi$) materially improves fit but
still misses the central tendency and the lower tail; only the further addition
of country-level random intercepts ($u_c$ with $\sigma_u \approx 0.72$ on the
logit scale) provides the best overall fit across DIC, central tendency,
variance, within-region heterogeneity, and country-level structure, although
residual lower-tail miscalibration remains. Mortality
remains negatively associated with success after pooling; incidence reverses to
a positive association once persistent country effects are absorbed; the
intercept implies a baseline success probability around 0.85. The full Bayesian
workflow — prior predictive checks, MCMC convergence diagnostics, observed-data
DIC, posterior predictive checks, and a 30/30/10-replicate parameter recovery
study — supports this recommendation. The frequentist comparison agrees on M1
and M2 sign and magnitude; the M3 GLMM analogue could not be fitted on the
local environment due to a Matrix/lme4 binary incompatibility, which we report
honestly as a non-fatal limitation of the bonus comparison.

---


# Course Requirements Traceability

| Guideline requirement | Where addressed |
|---|---|
| Meaningful title | Title page |
| Dataset and goals | Sections 3-4 |
| Dataset description and at least two alternative models | Sections 3-8 |
| Role of parameters and inferential goals | Sections 6-9 and 13 |
| Parameter recovery with simulated data | Section 12 |
| MCMC output illustration | Sections 10-11 |
| Bayesian estimation / hypothesis testing / prediction | Sections 13-14 |
| Model comparison through DIC | Section 15 |
| Ability to recover observed-data features | Section 14 |
| Bonus: model checking diagnostics | Sections 11 and 14 |
| Bonus: frequentist comparison | Section 16 |


# Reproducibility Appendix

## Software Versions

```{r session-info}
sessionInfo()
```

## Data Provenance

- Source: World Health Organization Global TB Programme
- Files downloaded: 2026-04-04
- Analysis window: 2012-2023
- Final sample: 1,862 country-years, 180 countries


## File Structure

- Main analysis script: `src/main.R`
- JAGS model files: `src/models/*.jags`
- Output tables: `src/outputs/tables/`
- Output figures: `src/outputs/figures/`

### How to regenerate the HTML report

The final HTML report is generated from:

`src/report/report.Rmd`

To regenerate:

```r
rmarkdown::render(
  "src/report/report.Rmd",
  output_format = "html_document",
  output_file = "report.html",
  output_dir = "src/report"
)
```

The full analysis source is `src/main.R`; JAGS model files are in `src/models/`.


# References

