---
title: "Bayesian Modeling of Cross-Country Tuberculosis Treatment Success"
subtitle: "A Fully Bayesian MCMC Analysis of WHO Data, 2012-2023"
author: "[Your Name]"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
bibliography: references.bib
csl: apa.csl
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "80%"
)

# Load packages
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)

# Set paths relative to report directory
PROJECT_ROOT <- normalizePath(file.path(getwd(), ".."))
TABLES_DIR <- file.path(PROJECT_ROOT, "outputs", "tables")
FIGURES_DIR <- file.path(PROJECT_ROOT, "outputs", "figures")
DIAGNOSTICS_DIR <- file.path(PROJECT_ROOT, "outputs", "diagnostics")
DATA_PROCESSED <- file.path(PROJECT_ROOT, "..", "data", "data_processed")
```

# Abstract

**Background:** Tuberculosis (TB) remains a major global health challenge. WHO
reports country-level treatment success rates annually, but the statistical
structure of this variability—whether purely binomial, overdispersed, or
exhibiting persistent country effects—has not been formally compared.
**Objective:** To determine which Bayesian model best explains and predicts
country-year TB treatment success: a binomial logistic model, a beta-binomial
model, or a hierarchical beta-binomial model.

**Methods:** We analyzed WHO TB data from 2012-2023 covering `r format(1862, big.mark=",")` 
country-years across 180 countries. Three Bayesian models of increasing complexity
were fitted via MCMC (JAGS) and compared using the Deviance Information Criterion
(DIC) and posterior predictive checks.

**Results:** [To be completed when Bayesian posteriors are available]

**Conclusion:** [To be completed when Bayesian posteriors are available]

**Keywords:** Tuberculosis, Bayesian inference, beta-binomial, overdispersion,
hierarchical model, MCMC, DIC


# Introduction & Research Gap

## Background

Tuberculosis (TB) remains one of the leading causes of death from a single
infectious agent worldwide. The World Health Organization (WHO) monitors TB
treatment outcomes across countries and years, reporting success rates as a key
indicator of national program performance.

## Research Gap

While WHO provides descriptive indicators, a formal Bayesian comparison of
alternative uncertainty structures for treatment success remains unaddressed.
The central inferential question is:

> Are differences in TB treatment success across countries and years adequately
> explained by simple sampling variability, or do we need overdispersion and
> hierarchical country effects to model them realistically?

## Research Question

**Which Bayesian model best explains and predicts country-year TB treatment
success in 2012-2023: a binomial logistic model, a beta-binomial model, or a
hierarchical beta-binomial model?**

This question is:

- **Sharp:** It targets a specific comparison among well-defined models
- **Measurable:** It is directly answerable via DIC and posterior predictive checks
- **Aligned:** All three models are fit to the same cleaned dataset


# Dataset & Analysis Goals

## Data Sources

```{r data-sources, results="asis"}
data_sources <- data.frame(
  File = c("TB_outcomes_2026-04-04.csv", "TB_burden_countries_2026-04-04.csv", 
           "TB_data_dictionary_2026-04-04.csv"),
  Role = c("Treatment outcomes (response)", "Epidemiological burden (predictors)", 
           "Variable definitions"),
  stringsAsFactors = FALSE
)

kable(data_sources, caption = "Data source files from WHO", booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))
```

## Unit of Analysis

One row = one country-year. The composite key is (iso3, year).

## Analysis Goals

1. Compare three Bayesian models using DIC and posterior predictive checks
2. Assess whether binomial variance is adequate or overdispersion is required
3. Quantify country-level heterogeneity in treatment success


# Data Construction & Cleaning

## Filtering Pipeline

```{r attrition-table}
attrition_path <- file.path(TABLES_DIR, "attrition_table.csv")
if (file.exists(attrition_path)) {
  attrition <- read.csv(attrition_path)
  kable(attrition, caption = "Sample attrition at each filtering step", 
        booktabs = TRUE) %>%
    kable_styling(latex_options = c("hold_position", "scale_down"))
}
```

## Final Sample

```{r final-sample}
snapshot_path <- file.path(TABLES_DIR, "final_sample_snapshot.csv")
if (file.exists(snapshot_path)) {
  snapshot <- read.csv(snapshot_path)
  kable(snapshot, caption = "Final sample characteristics", booktabs = TRUE) %>%
    kable_styling(latex_options = c("hold_position"))
}
```

## Attrition Flow

```{r attrition-flow, fig.cap="Attrition flow through filtering pipeline"}
attrition_fig <- file.path(FIGURES_DIR, "attrition_flow.png")
if (file.exists(attrition_fig)) knitr::include_graphics(attrition_fig)
```


# Exploratory Analysis

## Success Rate Distribution

```{r success-dist, fig.cap="Distribution of country-year treatment success rates"}
fig_path <- file.path(FIGURES_DIR, "success_rate_distribution.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

## Temporal Trends by Region

```{r temporal-trends, fig.cap="Temporal trends in treatment success by WHO region"}
fig_path <- file.path(FIGURES_DIR, "temporal_trend_by_region.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

## Overdispersion Evidence

```{r sd-ratio, fig.cap="Distribution of SD ratios (observed/expected binomial)"}
fig_path <- file.path(FIGURES_DIR, "country_sd_ratio_distribution.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

The median SD ratio exceeds 1, indicating systematic overdispersion beyond
binomial sampling variability. This motivates the beta-binomial extension.


# Model 1: Binomial Logistic

## Specification

$$Y_{it} \sim \text{Binomial}(n_{it}, p_{it})$$

$$\text{logit}(p_{it}) = \beta_0 + \mathbf{x}_{it}^\top \boldsymbol{\beta} + \gamma_{r[i]}$$

Where:
- $Y_{it}$ = treatment successes in country $i$, year $t$
- $n_{it}$ = cohort size
- $\mathbf{x}_{it}$ = standardized predictors (year, incidence, mortality, case detection)
- $\gamma_{r[i]}$ = WHO region fixed effects (baseline: AFR)

## Purpose

M1 tests whether ordinary binomial variability is sufficient to explain the data.


# Model 2: Beta-Binomial

## Specification

$$Y_{it} \sim \text{Beta-Binomial}(n_{it}, \mu_{it}, \phi)$$

$$\text{logit}(\mu_{it}) = \beta_0 + \mathbf{x}_{it}^\top \boldsymbol{\beta} + \gamma_{r[i]}$$

Where:
- $\phi$ = overdispersion parameter
- $\text{Var}(Y_{it}) = n_{it} \mu_{it}(1-\mu_{it}) \frac{\phi + n_{it}}{\phi + 1}$

## Purpose

M2 tests whether extra-binomial dispersion (captured by $\phi$) improves fit.
As $\phi \to \infty$, M2 reduces to M1.


# Model 3: Hierarchical Beta-Binomial

## Specification

$$Y_{it} \sim \text{Beta-Binomial}(n_{it}, \mu_{it}, \phi)$$

$$\text{logit}(\mu_{it}) = \beta_0 + \mathbf{x}_{it}^\top \boldsymbol{\beta} + \gamma_{r[i]} + u_i$$

$$u_i \sim N(0, \sigma_u^2)$$

Where:
- $u_i$ = country-specific random intercept
- $\sigma_u$ = between-country standard deviation

## Purpose

M3 tests whether persistent country heterogeneity (captured by $u_i$ and $\sigma_u$)
remains after controlling for observed covariates and overdispersion.


# Prior Specification

```{r prior-table}
prior_path <- file.path(TABLES_DIR, "prior_specification.csv")
if (file.exists(prior_path)) {
  priors <- read.csv(prior_path)
  kable(priors, caption = "Prior distributions for all model parameters", 
        booktabs = TRUE) %>%
    kable_styling(latex_options = c("hold_position"))
}
```

## Prior Predictive Checks

```{r prior-predictive, fig.cap="Prior predictive distributions showing plausibility of priors"}
fig_path <- file.path(FIGURES_DIR, "prior_predictive_combined.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

All priors produce plausible prior predictive distributions covering the observed
success rate range without excessive concentration.


# MCMC Implementation

*[This section will be completed when JAGS is available and Phase 8 is executed]*

## Configuration

- Chains: 4
- Adaptation: 1,000 iterations
- Burn-in: 4,000 iterations
- Post-burn-in: 8,000 iterations per chain
- Thinning: None
- Total posterior samples: 32,000


# MCMC Diagnostics

*[This section will be completed when JAGS is available and Phase 8 is executed]*

## Convergence Criteria

- R-hat < 1.05 for all parameters
- Effective sample size (ESS) > 400 for key parameters


# Parameter Recovery

*[This section will be completed when JAGS is available and Phase 11 is executed]*

## Recovery Study Design

- 50 simulated datasets per model
- True parameters: hand-chosen plausible values
- Evaluation metrics: bias, RMSE, 95% CI coverage


# Posterior Inference

*[This section will be completed when JAGS is available and Phase 9 is executed]*

## Expected Outputs

- Posterior summaries (mean, median, SD, 95% CI, 95% HPD)
- Directional probabilities for fixed effects
- Country random effect rankings (M3)


# Posterior Predictive Checks

*[This section will be completed when JAGS is available and Phase 10 is executed]*

## Test Quantities

1. Mean success rate (weighted and unweighted)
2. Variance of success rates
3. Count below 70% threshold
4. Within-region variance


# DIC Model Comparison

*[This section will be completed when JAGS is available and Phase 12 is executed]*

## DIC Interpretation

- $\Delta$DIC > 10: Strong evidence for lower-DIC model
- $\Delta$DIC 5-10: Moderate evidence
- $\Delta$DIC < 5: Interpret cautiously


# Frequentist Comparison

```{r frequentist-table}
freq_path <- file.path(TABLES_DIR, "bayesian_vs_frequentist_comparison.csv")
if (file.exists(freq_path)) {
  freq_comp <- read.csv(freq_path)
  kable(freq_comp, caption = "Bayesian vs. Frequentist coefficient comparison",
        booktabs = TRUE) %>%
    kable_styling(latex_options = c("hold_position", "scale_down"))
}
```

The frequentist comparison serves to validate Bayesian coefficient signs and
magnitudes. Agreement strengthens confidence in the findings.


# Sensitivity Analyses

```{r sensitivity-table}
sens_path <- file.path(TABLES_DIR, "sensitivity_summary.csv")
if (file.exists(sens_path)) {
  sens <- read.csv(sens_path)
  kable(sens, caption = "Summary of sensitivity analyses", booktabs = TRUE) %>%
    kable_styling(latex_options = c("hold_position", "scale_down"))
}
```

Five sensitivity analyses test robustness to:

1. Cohort threshold (cohort > 0 vs. cohort >= 50)
2. TB-HIV covariate inclusion
3. Overdispersion prior ($\phi$)
4. Country RE prior ($\sigma_u$)
5. Post-2021 definitions only


# Discussion

## Research Question Answer

*[To be completed when Bayesian posteriors are available]*

## Preferred Model Rationale

*[To be completed when Bayesian posteriors are available]*

## Limitations

1. **Ecological fallacy:** Country-level data cannot support individual-level
   causal claims.

2. **Reporting heterogeneity:** Countries differ in data quality, surveillance
   capacity, and conventions.

3. **Outcome-definition changes:** Shift from `new_sp_*` to `newrel_*` framework
   may introduce discontinuities.

4. **Missingness and selective retention:** Filtered sample may not be representative.

5. **Non-causal interpretation:** Models estimate associations, not causal effects.


# Conclusion

*[To be completed when Bayesian posteriors are available]*

This project developed a fully Bayesian comparison of three models for TB treatment
success. Based on DIC and posterior predictive checks, *[model recommendation]*.


# Reproducibility Appendix

## Software Versions

```{r session-info}
sessionInfo()
```

## Data Provenance

- Source: World Health Organization Global TB Programme
- Files downloaded: 2026-04-04
- Analysis window: 2012-2023
- Final sample: 1,862 country-years, 180 countries


## File Structure

- Main analysis script: `src/main.R`
- JAGS model files: `src/models/*.jags`
- Output tables: `src/outputs/tables/`
- Output figures: `src/outputs/figures/`


# References

