---
title: "Bayesian Modeling of Cross-Country Tuberculosis Treatment Success"
subtitle: "A Fully Bayesian MCMC Analysis of WHO Data, 2012–2023"
author: "Arman Feili"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 1
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: false
    theme: cosmo
    highlight: tango
    df_print: paged
    code_folding: hide
    self_contained: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "80%"
)

# Load packages
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)

# Set paths relative to report directory
PROJECT_ROOT <- normalizePath(file.path(getwd(), ".."))
TABLES_DIR <- file.path(PROJECT_ROOT, "outputs", "tables")
FIGURES_DIR <- file.path(PROJECT_ROOT, "outputs", "figures")
DIAGNOSTICS_DIR <- file.path(PROJECT_ROOT, "outputs", "diagnostics")
DATA_PROCESSED <- file.path(PROJECT_ROOT, "..", "data", "data_processed")

html_table <- function(x, caption = NULL, digits = 3) {
  knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    escape = TRUE
  ) |>
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed", "responsive"),
      full_width = TRUE,
      position = "left"
    )
}
```

<style>
body {
  font-family: Roboto, Arial, Helvetica, sans-serif;
  font-size: 16px;
  line-height: 1.7;
  color: #1f2933;
  background: #ffffff;
}

.main-container {
  max-width: 1200px !important;
}

.toc-content {
  padding-left: 28px;
  padding-right: 28px;
}

div.tocify {
  width: 20%;
  max-width: 260px;
  border: 1px solid #cccccc;
  border-radius: 6px;
  padding: 10px 8px;
  background: #f8fafc;
  max-height: 88vh !important;
  overflow-y: auto;
  box-shadow: 0 3px 10px rgba(15, 23, 42, 0.05);
}

#TOC {
  border: none;
  box-shadow: none;
  background: transparent;
}

.tocify .tocify-item > a,
.tocify .list-group-item {
  border: none;
  border-radius: 4px;
  margin-bottom: 1px;
  color: #334155;
  font-size: 14px;
  line-height: 1.35;
}

.tocify .list-group-item.active,
.tocify .tocify-item > a.tocify-focus {
  background-color: #2563eb;
  color: white;
}

h1, h2, h3, h4 {
  font-family: Roboto, Arial, Helvetica, sans-serif;
  font-weight: 700;
  color: #0f172a;
}

h1.title {
  font-size: 38px;
  font-weight: 800;
  line-height: 1.2;
}

h1 {
  font-size: 34px;
  margin-top: 40px;
  padding-bottom: 8px;
  border-bottom: 2px solid #e5e7eb;
}

h2 {
  font-size: 30px;
  margin-top: 30px;
}

h3 {
  font-size: 24px;
  margin-top: 24px;
}

h4 {
  font-size: 18px;
  margin-top: 18px;
}

#header {
  padding: 24px 28px;
  margin-bottom: 28px;
  border-radius: 12px;
  background: linear-gradient(135deg, #eff6ff 0%, #f8fafc 55%, #ffffff 100%);
  border: 1px solid #dbeafe;
  box-shadow: 0 6px 18px rgba(15, 23, 42, 0.08);
}

.title {
  font-size: 38px;
  font-weight: 700;
  color: #0f172a;
}

.subtitle {
  font-size: 21px;
  color: #475569;
}

table {
  font-size: 14px;
}

.table {
  width: 100%;
}

caption {
  color: #475569;
  font-weight: 600;
  margin-bottom: 8px;
}

img {
  border-radius: 10px;
  margin-top: 10px;
  margin-bottom: 10px;
  box-shadow: 0 4px 18px rgba(15, 23, 42, 0.08);
}

.figure-caption, em.figure-caption {
  display: block;
  color: #475569;
  font-size: 14px;
  margin-top: 6px;
  margin-bottom: 18px;
}

blockquote {
  border-left: 5px solid #2563eb;
  background: #eff6ff;
  padding: 12px 18px;
  border-radius: 8px;
}

code {
  color: #1f2933;
  background: #f1f5f9;
  padding: 1px 4px;
  border-radius: 4px;
}

pre {
  background: #ffffff;
  border-radius: 8px;
  border: 1px solid #e5e7eb;
}

pre code {
  background: transparent;
  padding: 0;
}

.tocify .tocify-subheader {
  display: none;
}

.tocify .tocify-subheader .tocify-item > a,
.tocify .tocify-subheader .list-group-item {
  padding-left: 18px;
  font-size: 13px;
}

.tocify .tocify-subheader .tocify-subheader .tocify-item > a,
.tocify .tocify-subheader .tocify-subheader .list-group-item {
  padding-left: 32px;
}

.tocify .toc-expanded > a {
  font-weight: 700;
}

@media (max-width: 991px) {
  div.tocify {
    width: 100%;
    max-width: none;
    max-height: none !important;
    margin-bottom: 16px;
  }

  .toc-content {
    margin-left: 0 !important;
    max-width: 100% !important;
    padding-left: 14px;
    padding-right: 14px;
  }
}

hr {
  border-top: 1px solid #e5e7eb;
  margin: 34px 0;
}
</style>

<script>
(function () {
  function initCollapsibleToc() {
    if (!window.jQuery) return false;
    var $ = window.jQuery;
    var $toc = $("#TOC");
    if (!$toc.length) return false;

    var $root = $toc.children("ul");
    if (!$root.length) return false;

    var $topItems = $root.children("li");
    if (!$topItems.length) return false;

    $topItems.each(function () {
      var $item = $(this);
      var $sub = $item.children("ul.tocify-subheader");
      if ($sub.length) {
        $item.addClass("toc-has-children");
        $sub.hide();
      }
    });

    function topItemFor($node) {
      if (!$node || !$node.length) return $();
      var $top = $node.closest("li");
      while ($top.length && !$top.parent().is($root)) {
        $top = $top.parent().closest("li");
      }
      return $top;
    }

    function expandTop($item, animate) {
      if (!$item.length) return;
      var $sub = $item.children("ul.tocify-subheader");
      if (!$sub.length) return;

      $topItems.not($item)
        .removeClass("toc-expanded")
        .children("ul.tocify-subheader")
        .stop(true, true)
        .slideUp(120);

      $item.addClass("toc-expanded");
      if (animate) {
        $sub.stop(true, true).slideDown(120);
      } else {
        $sub.show();
      }
    }

    function syncActiveBranch() {
      var $active = $toc.find("li.active, li.tocify-focus").first();
      if (!$active.length) return;
      var $top = topItemFor($active);
      if (!$top.length) return;
      expandTop($top, false);
      $active.parents("ul.tocify-subheader").show();
    }

    $topItems.children("a")
      .off("click.tocCollapse")
      .on("click.tocCollapse", function () {
        var $item = $(this).parent("li");
        var $sub = $item.children("ul.tocify-subheader");
        if (!$sub.length) return;

        if ($item.hasClass("toc-expanded")) {
          $item.removeClass("toc-expanded");
          $sub.stop(true, true).slideUp(120);
        } else {
          expandTop($item, true);
        }
      });

    syncActiveBranch();
    $(window)
      .off("scroll.tocCollapse hashchange.tocCollapse")
      .on("scroll.tocCollapse hashchange.tocCollapse", function () {
        syncActiveBranch();
      });

    return true;
  }

  function bootCollapsibleToc(triesLeft) {
    if (initCollapsibleToc()) return;
    if (triesLeft <= 0) return;
    setTimeout(function () {
      bootCollapsibleToc(triesLeft - 1);
    }, 250);
  }

  document.addEventListener("DOMContentLoaded", function () {
    bootCollapsibleToc(40);
  });
})();
</script>

# Overview

## Abstract

**Background:** Tuberculosis (TB) remains a major global health challenge. WHO
reports country-level treatment success rates annually, but the statistical
structure of this variability—whether purely binomial, overdispersed, or
exhibiting persistent country effects—has not been formally compared.
**Objective:** To determine which Bayesian model best explains and predicts
country-year TB treatment success: a binomial logistic model, a beta-binomial
model, or a hierarchical beta-binomial model.

**Methods:** We analyzed WHO TB data from 2012-2023 covering `r format(1862, big.mark=",")`
country-years across 180 countries. Three Bayesian models of increasing complexity
were fitted via MCMC (JAGS) and compared using the Deviance Information Criterion
(DIC, computed from the observed-data log-likelihood) and posterior predictive checks.
A reduced-design 30/30/10-replicate parameter recovery study supported inferential
calibration. Frequentist analogues were completed for M1 (binomial GLM) and M2
(beta-binomial via VGAM), while the M3 frequentist GLMM failed locally due to a
Matrix/lme4 binary incompatibility. Sensitivity material combined three executed
frequentist checks with two Bayesian prior-sensitivity specifications that were
prepared but not fully refitted.

**Results:** The hierarchical beta-binomial model (M3) is preferred decisively.
DIC is 24,940 for M3 versus 27,160 for M2 ($\Delta$DIC = 2,220) and 2,666,301 for
M1 ($\Delta$DIC $\approx$ 2.6 million), reflecting catastrophic underdispersion
of the binomial. Under M3 the posterior mean intercept is $\beta_0 = 1.77$
(95% CI 1.70–1.83); incidence has a positive posterior effect ($\beta = +0.28$,
95% CI 0.18–0.39, P > 0 = 1.000), mortality a negative effect ($\beta = -0.39$,
95% CI -0.47, -0.30, P < 0 = 1.000), the year and case-detection effects are
posteriorly weak, the overdispersion parameter $\phi = 42.9$ is finite, and the
country-effect SD $\sigma_u = 0.72$ (95% CI 0.64–0.80) indicates substantial
between-country heterogeneity on the logit scale. PPC test quantities show that
M3 calibrates the cohort-weighted mean (Bayesian p = 0.32) and approximately the
variance (p = 0.04), while M1 fails all variance-based checks (p = 0).

**Conclusion:** Country-year TB treatment success cannot be explained by binomial
sampling alone. Both extra-binomial overdispersion (M2) and persistent country
heterogeneity (M3) are required, and only M3 jointly captures the central
tendency, overall variance, and country-level structure of the WHO data,
although residual lower-tail and within-region variance miscalibration remains.

**Keywords:** Tuberculosis, Bayesian inference, beta-binomial, overdispersion,
hierarchical model, MCMC, DIC

> **Executive summary.** The hierarchical beta-binomial model is the preferred model. The binomial model severely underestimates variability, the beta-binomial improves overdispersion, and the hierarchical beta-binomial adds persistent country-level heterogeneity. The preferred model has the lowest observed-data DIC and the best posterior predictive behavior, although residual lower-tail and within-region variance miscalibration remains.

---

## Navigation Guide

This report is organized into eight major sections:

1. Overview
2. Data and EDA
3. Models and Priors
4. Computation and Recovery
5. Results and Model Checking
6. Robustness Checks
7. Discussion
8. Reproducibility

---


## Introduction

### Background

Tuberculosis (TB) remains one of the leading causes of death from a single
infectious agent worldwide. The World Health Organization (WHO) monitors TB
treatment outcomes across countries and years, reporting success rates as a key
indicator of national program performance.

### Research Gap

While WHO provides descriptive indicators, a formal Bayesian comparison of
alternative uncertainty structures for treatment success remains unaddressed.
The central inferential question is:

> Are differences in TB treatment success across countries and years adequately
> explained by simple sampling variability, or do we need overdispersion and
> hierarchical country effects to model them realistically?

## Objective and Research Question

### Objective

The objective is to identify the model that best explains and predicts
country-year TB treatment success while keeping interpretation transparent for
policy-oriented readers.

### Research Question

**Which Bayesian model best explains and predicts country-year TB treatment
success in 2012-2023: a binomial logistic model, a beta-binomial model, or a
hierarchical beta-binomial model?**

This question is:

- **Sharp:** It targets a specific comparison among well-defined models.
- **Measurable:** It is directly answerable via DIC and posterior predictive checks.
- **Aligned:** All three models are fit to the same cleaned dataset.


# Data and EDA

## Dataset Overview

### Data Sources

```{r data-sources, results="asis"}
data_sources <- data.frame(
  File = c("TB_outcomes_2026-04-04.csv", "TB_burden_countries_2026-04-04.csv", 
           "TB_data_dictionary_2026-04-04.csv"),
  Role = c("Treatment outcomes (response)", "Epidemiological burden (predictors)", 
           "Variable definitions"),
  stringsAsFactors = FALSE
)

html_table(data_sources, caption = "Data source files from WHO")
```

### Unit of Analysis

One row = one country-year. The composite key is (iso3, year).

### Analysis Goals

1. Compare three Bayesian models using DIC and posterior predictive checks
2. Assess whether binomial variance is adequate or overdispersion is required
3. Quantify country-level heterogeneity in treatment success

---


## Data Preparation

### Filtering Pipeline

```{r attrition-table}
attrition_path <- file.path(TABLES_DIR, "attrition_table.csv")
if (file.exists(attrition_path)) {
  attrition <- read.csv(attrition_path)
  html_table(attrition, caption = "Sample attrition at each filtering step")
}
```

### Final Sample

```{r final-sample}
snapshot_path <- file.path(TABLES_DIR, "final_sample_snapshot.csv")
if (file.exists(snapshot_path)) {
  snapshot <- read.csv(snapshot_path)
  html_table(snapshot, caption = "Final sample characteristics")
}
```

### Predictor Screening and Collinearity

```{r predictor-screening, results="asis"}
corr_path <- file.path(TABLES_DIR, "predictor_correlations.csv")
vif_path <- file.path(TABLES_DIR, "predictor_vif.csv")

max_abs_corr <- NA_real_
max_vif <- NA_real_

if (file.exists(corr_path)) {
  corr <- read.csv(corr_path, stringsAsFactors = FALSE)
  html_table(corr, caption = "Pairwise correlations among continuous predictors", digits = 3)

  corr_num <- suppressWarnings(as.numeric(corr$abs_correlation))
  if (length(corr_num) && any(is.finite(corr_num))) {
    max_abs_corr <- max(corr_num, na.rm = TRUE)
  }
}

if (file.exists(vif_path)) {
  vif <- read.csv(vif_path, stringsAsFactors = FALSE)
  html_table(vif, caption = "VIF-style collinearity diagnostics", digits = 3)

  vif_num <- suppressWarnings(as.numeric(vif$VIF))
  if (length(vif_num) && any(is.finite(vif_num))) {
    max_vif <- max(vif_num, na.rm = TRUE)
  }
}

corr_fig_path <- file.path(FIGURES_DIR, "predictor_correlation_matrix.png")
if (file.exists(corr_fig_path)) knitr::include_graphics(corr_fig_path)
```

*Figure: Correlation matrix of standardized continuous predictors used in the primary Bayesian models.*

The main continuous predictors were screened for collinearity before fitting the
Bayesian models. The largest pairwise correlation is between incidence and
mortality. In the current diagnostics, max $|r|$ is `r ifelse(is.finite(max_abs_corr), sprintf('%.3f', max_abs_corr), 'not available')`,
and max VIF is `r ifelse(is.finite(max_vif), sprintf('%.2f', max_vif), 'not available')`.
Both remain below the pre-specified exclusion threshold and below the common VIF
warning level of 5, so year, incidence, mortality, and case detection were
retained together in all three primary Bayesian models.

### Attrition Flow

```{r attrition-flow}
attrition_fig <- file.path(FIGURES_DIR, "attrition_flow.png")
if (file.exists(attrition_fig)) knitr::include_graphics(attrition_fig)
```

*Figure: Attrition flow through filtering pipeline.*

Interpretation: This flow summarizes how the merged raw dataset is reduced by
pre-specified eligibility and quality filters into the locked analytic sample
used for all primary Bayesian models.

### Stepwise Sample Retention Across Filtering Stages

```{r attrition-waterfall}
retention_fig <- file.path(FIGURES_DIR, "attrition_waterfall.png")
if (file.exists(retention_fig)) knitr::include_graphics(retention_fig)
```

*Figure: Remaining country-years after each preprocessing step.*

Interpretation: This step-count view shows where the largest sample reductions
occur and confirms that the final analysis uses a transparent, reproducible,
locked subset of the original merged panel.

### Region-Year Coverage After Filtering

```{r region-year-retention-heatmap}
retention_heatmap <- file.path(FIGURES_DIR, "region_year_retention_heatmap.png")
if (file.exists(retention_heatmap)) knitr::include_graphics(retention_heatmap)
```

*Figure: Retained country-year counts by WHO region and year after preprocessing.*

Interpretation: Despite filtering, the final sample still spans all WHO regions
and the full 2012–2023 window, supporting region-aware temporal comparisons in
the downstream Bayesian models.

---


## Exploratory Analysis

### Success-Rate Distribution by Region

```{r success-dist}
fig_path <- file.path(FIGURES_DIR, "success_rate_by_region_density.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

*Figure: Regional density distributions of country-year treatment success rates.*

Interpretation: Most country-years cluster at relatively high success rates,
but there is clear between-region heterogeneity and a persistent lower tail,
which motivates models that allow richer variance and country-level structure.

### Temporal Trends by Region

```{r temporal-trends}
fig_path <- file.path(FIGURES_DIR, "temporal_trend_faceted.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

*Figure: Faceted temporal trends in treatment success by WHO region.*

Interpretation: Regional trajectories are not identical over time. Some regions
show stable high performance, while others show slower improvement and wider
dispersion, motivating region adjustment and country-level heterogeneity.

---


# Models and Priors

## Bayesian Modeling Roadmap

The model sequence is intentionally incremental.

| Model | Main idea | What it tests | Expected limitation |
|---|---|---|---|
| M1 (Binomial logistic) | Fixed effects with binomial sampling variation only | Whether ordinary binomial variability is enough for country-year TB success data | Typically underestimates variance when data are overdispersed |
| M2 (Beta-binomial) | Adds overdispersion parameter $\phi$ | Whether extra-binomial variability materially improves calibration and fit | Does not explicitly model persistent country-specific heterogeneity |
| M3 (Hierarchical beta-binomial) | Adds country random intercepts $u_i$ and SD $\sigma_u$ | Whether persistent country-level effects remain after covariate adjustment and overdispersion | More complex estimation; can still miss extreme lower-tail behavior |

This table summarizes why the models are compared in this order and how each
step adds one new component of data-generating realism.

---


## Model 1 — Binomial Logistic Baseline

### Specification

$$Y_{it} \sim \text{Binomial}(n_{it}, p_{it})$$

$$\text{logit}(p_{it}) = \beta_0 + \mathbf{x}_{it}^\top \boldsymbol{\beta} + \gamma_{r[i]}$$

Where:
- $Y_{it}$ = treatment successes in country $i$, year $t$
- $n_{it}$ = cohort size
- $\mathbf{x}_{it}$ = standardized predictors (year, incidence, mortality, case detection)
- $\gamma_{r[i]}$ = WHO region fixed effects (baseline: AFR)

### Purpose

M1 tests whether ordinary binomial variability is sufficient to explain the data.

### In Simple Terms

Model 1 asks: "If every country-year followed the same basic success-rate
equation and differences were only due to random binomial sampling, would that
already be enough?" It is the simplest benchmark and the model we expect to
fail first if real-world variability is richer.


## Model 2 — Beta-Binomial Overdispersion Model

### Specification

$$Y_{it} \sim \text{Beta-Binomial}(n_{it}, \mu_{it}, \phi)$$

$$\text{logit}(\mu_{it}) = \beta_0 + \mathbf{x}_{it}^\top \boldsymbol{\beta} + \gamma_{r[i]}$$

Where:
- $\phi$ = overdispersion parameter

\[
\mathrm{Var}(Y_{it}\mid \mu_{it},\phi)
=
n_{it}\mu_{it}(1-\mu_{it})
\frac{n_{it}+\phi}{1+\phi}.
\]

Compared with the binomial variance, the beta-binomial inflates the variance by
the factor $(n_{it}+\phi)/(1+\phi)$.

### Purpose

M2 tests whether extra-binomial dispersion (captured by $\phi$) improves fit.
As $\phi \to \infty$, M2 reduces to M1.

### In Simple Terms

Model 2 says: "Maybe countries are more variable than the binomial model
allows." The parameter $\phi$ controls that extra variability. If M2 fits much
better than M1, overdispersion is real and cannot be ignored.


## Model 3 — Hierarchical Beta-Binomial Country-Effects Model

### Specification

$$Y_{it} \sim \text{Beta-Binomial}(n_{it}, \mu_{it}, \phi)$$

$$\text{logit}(\mu_{it}) = \beta_0 + \mathbf{x}_{it}^\top \boldsymbol{\beta} + \gamma_{r[i]} + u_i$$

$$u_i \sim N(0, \sigma_u^2)$$

Where:
- $u_i$ = country-specific random intercept
- $\sigma_u$ = between-country standard deviation

### Purpose

M3 tests whether persistent country heterogeneity (captured by $u_i$ and $\sigma_u$)
remains after controlling for observed covariates and overdispersion.

### In Simple Terms

Model 3 says: "Even after controlling for measured predictors, countries may
have stable, unmeasured differences in treatment performance." The random effect
$u_i$ captures each country's persistent offset, and $\sigma_u$ summarizes how
large those cross-country differences are overall.

---


## Prior Specification and Prior Predictive Checks

```{r prior-table}
prior_path <- file.path(TABLES_DIR, "prior_specification.csv")
if (file.exists(prior_path)) {
  priors <- read.csv(prior_path, stringsAsFactors = FALSE)
  html_table(priors, caption = "Prior distributions for all model parameters")
}
```

### Prior Predictive Checks

```{r prior-predictive}
fig_path <- file.path(FIGURES_DIR, "prior_predictive_combined.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

*Figure: Prior predictive distributions showing plausibility of priors.*

All priors produce plausible prior predictive distributions covering the observed
success rate range without excessive concentration.

---


# Computation and Recovery

## MCMC Implementation

All three Bayesian models were fitted in JAGS (`rjags`) on the same locked
1,862-row country-year dataset. Settings:

- Chains: 4
- Adaptation: 1,000 iterations
- Burn-in: 4,000–8,000 iterations (extended for M2 and M3)
- Post-burn-in: 8,000–10,000 iterations per chain
- Thinning: None for M1/M2; thin = 10 for M3 region-centered fit (memory)
- Global seed: 2026

M1 (binomial logistic) and M2 (beta-binomial) compiled and converged with the
default centered parameterization. M3 required substantial remediation. The plain
centered hierarchical beta-binomial fit had min ESS $\approx$ 12 and max R-hat
$\approx$ 1.16 on $\beta_0$ and $\gamma$ even with very long runs, an
identifiability problem caused by an additive ridge between $\beta_0$, $\gamma_r$,
and the within-region mean of $u_c$. Three escalating non-centered budgets did
not resolve it. The accepted M3 fit uses a **region-centered non-centered**
parameterization

$$ u_c = \sigma_u \, (z_c - \bar{z}_{r(c)}), \quad z_c \sim N(0,1), $$

which enforces $\sum_{c \in r} u_c = 0$ within every WHO region. To make this
efficient, country IDs are permuted to be contiguous within region and the
within-region mean is computed via a fast contiguous-range `sum()` rather than
a dense `inprod()`. Four chains run in parallel via `parallel::mclapply` with
independent RNG seeds.


## MCMC Diagnostics

```{r mcmc-diag-summary}
diag_path <- file.path(TABLES_DIR, "mcmc_diagnostics_summary.csv")
if (file.exists(diag_path)) {
  diag <- read.csv(diag_path)
  html_table(diag, caption = "MCMC diagnostics summary on key globals")
}
```

```{r compact-mcmc-summary, results="asis"}
diag_file_candidates <- list(
  M1 = c(
    file.path(DIAGNOSTICS_DIR, "m1_extended_diagnostics_summary.csv"),
    file.path(DIAGNOSTICS_DIR, "m1_diagnostics_summary.csv")
  ),
  M2 = c(
    file.path(DIAGNOSTICS_DIR, "m2_extended_diagnostics_summary.csv"),
    file.path(DIAGNOSTICS_DIR, "m2_diagnostics_summary.csv")
  ),
  M3 = c(
    file.path(DIAGNOSTICS_DIR, "m3_regioncentered_full_diagnostics_summary.csv"),
    file.path(DIAGNOSTICS_DIR, "m3_diagnostics_summary.csv")
  )
)

read_first_existing <- function(paths) {
  hit <- paths[file.exists(paths)][1]
  if (is.na(hit)) return(NULL)
  x <- read.csv(hit, stringsAsFactors = FALSE)
  attr(x, "source_file") <- basename(hit)
  x
}

extract_diag_summary <- function(model_name, paths) {
  x <- read_first_existing(paths)
  if (is.null(x)) {
    return(data.frame(
      Model = model_name,
      Source = NA_character_,
      Max_Rhat = NA_real_,
      Min_ESS = NA_real_,
      Status = "Missing diagnostics file",
      stringsAsFactors = FALSE
    ))
  }

  nms <- names(x)
  lower <- tolower(nms)

  param_col <- nms[grepl("^parameter$|^param$|name", lower)][1]
  if (is.na(param_col)) {
    return(data.frame(
      Model = model_name,
      Source = attr(x, "source_file"),
      Max_Rhat = NA_real_,
      Min_ESS = NA_real_,
      Status = "CHECK",
      stringsAsFactors = FALSE
    ))
  }

  rhat_col <- nms[grepl("^rhat$|^r_hat$|^r\\.hat$", lower)][1]
  if (is.na(rhat_col)) {
    rhat_col <- nms[grepl("rhat|r_hat|r\\.hat", lower) & !grepl("flag", lower)][1]
  }

  ess_col <- nms[grepl("^ess$|^n_eff$|^neff$|^effective_sample_size$", lower)][1]
  if (is.na(ess_col)) {
    ess_col <- nms[grepl("ess|effective", lower) & !grepl("flag", lower)][1]
  }

  p <- as.character(x[[param_col]])
  keep <- startsWith(p, "beta[") |
    p %in% c("beta0", "beta_0", "phi", "sigma_u") |
    grepl("^gamma\\[[2-9][0-9]*\\]$", p, perl = TRUE)

  if ("nonvarying" %in% nms) {
    nv <- as.logical(x[["nonvarying"]])
    nv[is.na(nv)] <- FALSE
    keep <- keep & !nv
  }

  xk <- x[keep, , drop = FALSE]

  rhat_vals <- if (!is.na(rhat_col)) suppressWarnings(as.numeric(xk[[rhat_col]])) else numeric(0)
  ess_vals <- if (!is.na(ess_col)) suppressWarnings(as.numeric(xk[[ess_col]])) else numeric(0)

  rhat_vals <- rhat_vals[is.finite(rhat_vals)]
  ess_vals <- ess_vals[is.finite(ess_vals) & ess_vals > 0]

  max_rhat <- if (length(rhat_vals)) max(rhat_vals, na.rm = TRUE) else NA_real_
  min_ess <- if (length(ess_vals)) min(ess_vals, na.rm = TRUE) else NA_real_

  status <- if (is.finite(max_rhat) && is.finite(min_ess) && max_rhat <= 1.05 && min_ess >= 400) {
    "PASS"
  } else {
    "CHECK"
  }

  data.frame(
    Model = model_name,
    Source = attr(x, "source_file"),
    Max_Rhat = max_rhat,
    Min_ESS = min_ess,
    Status = status,
    stringsAsFactors = FALSE
  )
}

diag_summary <- do.call(
  rbind,
  Map(extract_diag_summary, names(diag_file_candidates), diag_file_candidates)
)
rownames(diag_summary) <- NULL

html_table(diag_summary, caption = "Compact MCMC diagnostics summary for accepted fits", digits = 3)
```

The compact diagnostics table reports the final promoted fit for each model.
The primary acceptance rule was max R-hat $\leq$ 1.05 and min ESS $\geq$ 400 for
key global parameters. M3 is the hardest model computationally, but the
accepted region-centered non-centered fit satisfies the diagnostic threshold and
is therefore used for posterior inference, PPC, and DIC.

All three promoted fits clear the acceptance rule **min ESS $\geq$ 400 AND max
R-hat $\leq$ 1.05** on key globals ($\beta_0, \beta_{1:4}, \gamma_{2:6}, \phi, \sigma_u$).
M3 reaches min ESS = 407.1 and max R-hat = 1.0072 under the region-centered
parameterization. Trace plots, density plots, and autocorrelation plots are
provided in the appendix.

```{r m3-trace}
fig_candidates <- c(
  file.path(FIGURES_DIR, "m3_regioncentered_full_trace_density_core.png"),
  file.path(FIGURES_DIR, "m3_trace_density_core.png")
)
hit <- fig_candidates[file.exists(fig_candidates)][1]
if (!is.na(hit)) knitr::include_graphics(hit)
```

*Figure: Final accepted M3 trace and density diagnostics (region-centered non-centered fit).*

Interpretation: Chains overlap well and explore the same posterior region
without persistent drift. Together with ESS/R-hat thresholds, this supports
using these draws for inference rather than treating them as preliminary.

### Supplementary M1/M2 trace-density figures

**M1 promoted fit**

Preferred file: `m1_extended_trace_density_core.png` (fallback: `m1_trace_density_core.png`).

```{r m1-supp-trace-density, results='asis'}
m1_fig <- file.path(FIGURES_DIR, "m1_extended_trace_density_core.png")
if (!file.exists(m1_fig)) {
  m1_fig <- file.path(FIGURES_DIR, "m1_trace_density_core.png")
}
if (file.exists(m1_fig)) {
  knitr::include_graphics(m1_fig)
} else {
  cat("Figure not found: m1_extended_trace_density_core.png (fallback: m1_trace_density_core.png)\n\n")
}
```

**M2 promoted fit**

Preferred file: `m2_extended_trace_density_core.png` (fallback: `m2_trace_density_core.png`).

```{r m2-supp-trace-density, results='asis'}
m2_fig <- file.path(FIGURES_DIR, "m2_extended_trace_density_core.png")
if (!file.exists(m2_fig)) {
  m2_fig <- file.path(FIGURES_DIR, "m2_trace_density_core.png")
}
if (file.exists(m2_fig)) {
  knitr::include_graphics(m2_fig)
} else {
  cat("Figure not found: m2_extended_trace_density_core.png (fallback: m2_trace_density_core.png)\n\n")
}
```

---


## Parameter Recovery with Simulated Data

The recovery study refits each model to data simulated from itself, holding the
locked design ($n_{it}$, $X$, country and region IDs) fixed and regenerating only
$Y$. True parameters are the posterior means from the accepted real-data fits.

```{r recovery-failure}
fail_path <- file.path(TABLES_DIR, "recovery_failure_summary.csv")
if (file.exists(fail_path)) {
  fail <- read.csv(fail_path)
  html_table(fail, caption = "Recovery replicate counts and runtime", digits = 3)
}
```

The originally targeted design was 50 replicates × 3 models. The implementation
ran 30 successful replicates for M1, 30 for M2, and 10 for M3, with a 0% failure
rate for executed replicates. This is a documented computational accommodation,
not a methodological failure: M2/M3, especially M3, were substantially more
expensive per replicate than M1, so the project used a reduced but completed
design for feasible turnaround within the final run window.
The 0% value refers specifically to the predefined execution/failure rule for
recovery replicates.

```{r recovery-executive-summary}
fail_path <- file.path(TABLES_DIR, "recovery_failure_summary.csv")
perf_path <- file.path(TABLES_DIR, "recovery_performance.csv")
if (file.exists(fail_path) && file.exists(perf_path)) {
  fail <- read.csv(fail_path, stringsAsFactors = FALSE)
  perf <- read.csv(perf_path, stringsAsFactors = FALSE)

  cov_by_model <- aggregate(coverage_95 ~ model, data = perf, FUN = mean)
  names(cov_by_model)[2] <- "mean_coverage_95"

  recov_summary <- merge(
    fail[, c("model", "n_ok", "total_reps", "failure_rate")],
    cov_by_model,
    by = "model",
    all.x = TRUE
  )
  recov_summary$executed <- paste0(recov_summary$n_ok, "/", recov_summary$total_reps)
  recov_summary$failure_rate_pct <- sprintf("%.0f%%", 100 * recov_summary$failure_rate)

  html_table(
    recov_summary[, c("model", "executed", "failure_rate_pct", "mean_coverage_95")],
    caption = "Executed recovery design summary",
    digits = 3
  )
}
```

Interpretation: Recovery here is a calibration check. If true simulated values
are repeatedly recovered with near-nominal coverage, then the inference pipeline
is behaving credibly under known ground truth.

### Full recovery performance table

```{r recovery-perf}
perf_path <- file.path(TABLES_DIR, "recovery_performance.csv")
if (file.exists(perf_path)) {
  perf <- read.csv(perf_path)
  html_table(perf, caption = "Recovery performance (bias, RMSE, 95\\% CI coverage)", digits = 4)
}
```

```{r recovery-coverage}
fig_path <- file.path(FIGURES_DIR, "recovery_coverage_plot.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

*Figure: 95% credible-interval coverage of true parameter values.*

Mean coverage across key globals is 0.93 for M1 (10 parameters), 0.96 for M2 (11
parameters, including $\phi$), and 0.97 for M3 (12 parameters, including $\phi$
and $\sigma_u$). Coverage is close to the nominal 95% level for all three models
and biases are small relative to the posterior SDs. M3 mixing on $\beta_0$ and
$\gamma$ is intrinsically harder per replicate than in the main fit (each
replicate has a freshly simulated $Y$); some replicate-level R-hats therefore
exceed the main-fit threshold even though the recovery target is captured.
Because only 10 M3 replicates were computationally feasible, precision for M3
recovery summaries is lower than for M1/M2, but the overall calibration pattern
remains supportive of inferential credibility.
Some M3 parameter-level recovery diagnostics are weaker than the stricter
main-fit thresholds, so the recovery results are interpreted as supportive
calibration evidence rather than a stronger precision claim.

Interpretation: Coverage near 0.95 across all models indicates that intervals
are well calibrated on average, with M3 showing acceptable performance despite
its higher computational complexity.

---


**How to read posterior inference.** Coefficients are on the logit scale.
Positive values imply higher predicted treatment success (all else equal),
negative values imply lower predicted success. Credible intervals quantify
posterior uncertainty around each effect. Directional probabilities such as
`P > 0` or `P < 0` summarize posterior support for the effect direction.


# Results and Model Checking

## Posterior Inference

```{r posterior-m3}
post_path <- file.path(TABLES_DIR, "posterior_summary_m3.csv")
if (file.exists(post_path)) {
  post <- read.csv(post_path)
  html_table(post, caption = "M3 posterior summaries (preferred model)", digits = 3)
}
```

Interpretation: This table is the core inferential output for the preferred
model. Signs, interval widths, and posterior tail probabilities should be read
together rather than from point estimates alone.

```{r hyp-tests}
hyp_path <- file.path(TABLES_DIR, "hypothesis_tests_summary.csv")
if (file.exists(hyp_path)) {
  hyp <- read.csv(hyp_path)
  html_table(hyp, caption = "Directional posterior probabilities by model", digits = 3)
}
```

Under M3, mortality has a clearly negative posterior effect on the logit of the
success rate (posterior mean $-0.39$, 95% credible interval $[-0.47, -0.30]$,
posterior probability of being negative $= 1.000$). The posterior mean for
incidence is positive ($+0.28$, 95% CI $[0.18, 0.39]$, P>0 = 1.000) — opposite
to the M1 sign. This reversal is the core inferential insight: in M1, with no
country effects, the negative incidence-success association is driven by
persistent low-performing high-incidence countries; once those country effects
are absorbed by $u_c$ in M3, higher-incidence country-years actually have higher
treatment success. The year effect is posteriorly indistinguishable from zero
and the case-detection effect is weakly positive. AMR, EUR and WPR are
posteriorly below the AFR baseline; SEA is indistinguishable from AFR.
$\phi = 42.9$ is well bounded (95% CI 39.6–46.3) — finite, so overdispersion is
present — and $\sigma_u = 0.72$ (95% CI 0.64–0.80) is substantial.

```{r country-re}
fig_path <- file.path(FIGURES_DIR, "country_re_caterpillar_plot.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

*Figure: Country random-effect caterpillar plot (M3).*

Interpretation: Countries with intervals clearly above zero are persistently
high-performing relative to the adjusted global baseline, while those below zero
are persistently low-performing. This is descriptive benchmarking, not a causal
ranking.

### Highest and Lowest Country Random Effects

**Top 10 positive and top 10 negative adjusted country random effects**

```{r country-re-top-bottom, results="asis"}
top_pos_path <- file.path(TABLES_DIR, "country_re_top10_positive.csv")
top_neg_path <- file.path(TABLES_DIR, "country_re_top10_negative.csv")

keep_cols <- c("iso3", "mean", "ci_lower", "ci_upper", "g_whoregion")

if (file.exists(top_pos_path)) {
  top_pos <- read.csv(top_pos_path, stringsAsFactors = FALSE)
  cols <- intersect(keep_cols, names(top_pos))
  if (length(cols)) top_pos <- top_pos[cols]
  html_table(top_pos, caption = "Top 10 positive country random effects (M3)", digits = 3)
}

if (file.exists(top_neg_path)) {
  top_neg <- read.csv(top_neg_path, stringsAsFactors = FALSE)
  cols <- intersect(keep_cols, names(top_neg))
  if (length(cols)) top_neg <- top_neg[cols]
  html_table(top_neg, caption = "Top 10 negative country random effects (M3)", digits = 3)
}
```

These tables summarize the strongest positive and negative country-level
residual effects under M3. They should be read as descriptive adjusted country
offsets, not causal rankings. Positive values indicate countries with
persistently higher treatment success than expected after covariate and region
adjustment; negative values indicate persistently lower adjusted performance.


## Posterior Predictive Checks

### Recovering Observed Data Features

Detailed PPC quantities are provided below.

### Full posterior predictive check table

```{r ppc-table}
ppc_path <- file.path(TABLES_DIR, "ppc_summary_table.csv")
if (file.exists(ppc_path)) {
  ppc <- read.csv(ppc_path)
  html_table(ppc, caption = "PPC test quantities and Bayesian p-values", digits = 4)
}
```

Interpretation: The PPC table is read by checking whether Bayesian p-values are
close to 0.5 (well calibrated) or near 0/1 (systematic misfit). It complements
parameter-level summaries by testing whether models reproduce data-level
features we care about.

The six posterior predictive test quantities give a clear picture. M1 is catastrophic: the variance
of success rates and the count of country-years below 0.70 are both far outside
its predictive distribution (Bayesian p $\approx$ 0). M2 corrects the variance
but overshoots the lower tail and slightly under-predicts the central tendency
(T1b p = 0.005). M3 calibrates the cohort-weighted mean (T1b p = 0.32) and
approximately captures the variance (T2 p = 0.04), but residual miscalibration
remains in the lower tail (T3 p = 1.0) and within-region variance checks
(T4a/T4b p-values also low). Overall, M3 is the best-performing model among the
three candidates, not a perfect data-generating model.

```{r ppc-densities}
fig_path <- file.path(FIGURES_DIR, "ppc_m3_density_overlay.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

*Figure: PPC density overlays for M3.*

Interpretation: Under the preferred model, replicated datasets broadly recover
the observed success-rate distribution shape, indicating improved global fit
relative to simpler alternatives.

```{r ppc-variance-model-compare}
fig_path <- file.path(FIGURES_DIR, "ppc_model_comparison_variance.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

*Figure: Posterior predictive comparison of variance recovery across M1, M2, and M3.*

Interpretation: This figure visualizes the main PPC failure of M1: the binomial
model cannot reproduce the observed variance. M2 and M3 are much closer, with
M3 preferred overall because it also captures persistent country-level
structure.

```{r ppc-m3-cohort-calibration}
fig_path <- file.path(FIGURES_DIR, "ppc_m3_cohort_calibration.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

*Figure: M3 posterior predictive calibration across cohort-size strata.*

Interpretation: Calibration is acceptable across cohort-size strata, indicating
that the preferred model is not fitting only one cohort-size regime but tracks
observed behavior across small and large cohorts.

### Per-model PPC test-statistic visual diagnostics

**M1 PPC test statistics**

Source file: ppc_m1_test_statistics.png.

```{r ppc-model-test-stats-m1, results='asis'}
fp <- file.path(FIGURES_DIR, "ppc_m1_test_statistics.png")
if (file.exists(fp)) {
  knitr::include_graphics(fp)
} else {
  cat("Figure not found: ppc_m1_test_statistics.png\n\n")
}
```

**M2 PPC test statistics**

Source file: ppc_m2_test_statistics.png.

```{r ppc-model-test-stats-m2, results='asis'}
fp <- file.path(FIGURES_DIR, "ppc_m2_test_statistics.png")
if (file.exists(fp)) {
  knitr::include_graphics(fp)
} else {
  cat("Figure not found: ppc_m2_test_statistics.png\n\n")
}
```

**M3 PPC test statistics**

Source file: ppc_m3_test_statistics.png.

```{r ppc-model-test-stats-m3, results='asis'}
fp <- file.path(FIGURES_DIR, "ppc_m3_test_statistics.png")
if (file.exists(fp)) {
  knitr::include_graphics(fp)
} else {
  cat("Figure not found: ppc_m3_test_statistics.png\n\n")
}
```

### Supplementary EDA and Diagnostic Visuals

**Cohort distribution by region**

Source file: cohort_distribution_by_region.png.

```{r supp-cohort-distribution-by-region, results='asis'}
fp <- file.path(FIGURES_DIR, "cohort_distribution_by_region.png")
if (file.exists(fp)) {
  knitr::include_graphics(fp)
} else {
  cat("Figure not found: cohort_distribution_by_region.png\n\n")
}
```

*Caption: Regional cohort-size distributions for sample composition context.*

**Bivariate plots (combined)**

Source file: bivariate_plots_combined.png.

```{r supp-bivariate-plots-combined, results='asis'}
fp <- file.path(FIGURES_DIR, "bivariate_plots_combined.png")
if (file.exists(fp)) {
  knitr::include_graphics(fp)
} else {
  cat("Figure not found: bivariate_plots_combined.png\n\n")
}
```

*Caption: Supplementary bivariate relationships among key predictors and outcomes.*

**Region-year success heatmap**

Source file: region_year_success_heatmap.png.

```{r supp-region-year-success-heatmap, results='asis'}
fp <- file.path(FIGURES_DIR, "region_year_success_heatmap.png")
if (file.exists(fp)) {
  knitr::include_graphics(fp)
} else {
  cat("Figure not found: region_year_success_heatmap.png\n\n")
}
```

*Caption: Regional temporal structure in observed treatment success rates.*

**Country SD ratio distribution**

Source file: country_sd_ratio_distribution.png.

```{r supp-country-sd-ratio-distribution, results='asis'}
fp <- file.path(FIGURES_DIR, "country_sd_ratio_distribution.png")
if (file.exists(fp)) {
  knitr::include_graphics(fp)
} else {
  cat("Figure not found: country_sd_ratio_distribution.png\n\n")
}
```

*Caption: Distribution of country-level variability relative to aggregate spread.*

**M1 autocorrelation diagnostics**

Source file: m1_autocorr_core.png.

```{r supp-m1-autocorr-core, results='asis'}
fp <- file.path(FIGURES_DIR, "m1_autocorr_core.png")
if (file.exists(fp)) {
  knitr::include_graphics(fp)
} else {
  cat("Figure not found: m1_autocorr_core.png\n\n")
}
```

*Caption: Supplementary autocorrelation diagnostics for M1 chains.*

**M2 autocorrelation diagnostics**

Source file: m2_autocorr_core.png.

```{r supp-m2-autocorr-core, results='asis'}
fp <- file.path(FIGURES_DIR, "m2_autocorr_core.png")
if (file.exists(fp)) {
  knitr::include_graphics(fp)
} else {
  cat("Figure not found: m2_autocorr_core.png\n\n")
}
```

*Caption: Supplementary autocorrelation diagnostics for M2 chains.*

**M3 autocorrelation diagnostics (region-centered full)**

Source file: m3_regioncentered_full_autocorr_core.png.

```{r supp-m3-regioncentered-full-autocorr-core, results='asis'}
fp <- file.path(FIGURES_DIR, "m3_regioncentered_full_autocorr_core.png")
if (file.exists(fp)) {
  knitr::include_graphics(fp)
} else {
  cat("Figure not found: m3_regioncentered_full_autocorr_core.png\n\n")
}
```

*Caption: Supplementary autocorrelation diagnostics for the accepted M3 fit.*

**M1 cohort calibration PPC**

Source file: ppc_m1_cohort_calibration.png.

```{r supp-ppc-m1-cohort-calibration, results='asis'}
fp <- file.path(FIGURES_DIR, "ppc_m1_cohort_calibration.png")
if (file.exists(fp)) {
  knitr::include_graphics(fp)
} else {
  cat("Figure not found: ppc_m1_cohort_calibration.png\n\n")
}
```

*Caption: Supplementary cohort-stratified PPC calibration for M1.*

**M2 cohort calibration PPC**

Source file: ppc_m2_cohort_calibration.png.

```{r supp-ppc-m2-cohort-calibration, results='asis'}
fp <- file.path(FIGURES_DIR, "ppc_m2_cohort_calibration.png")
if (file.exists(fp)) {
  knitr::include_graphics(fp)
} else {
  cat("Figure not found: ppc_m2_cohort_calibration.png\n\n")
}
```

*Caption: Supplementary cohort-stratified PPC calibration for M2.*


## Model Comparison Using Observed-Data DIC

```{r dic-table}
dic_path <- file.path(TABLES_DIR, "dic_comparison_table.csv")
if (file.exists(dic_path)) {
  dic <- read.csv(dic_path)
  html_table(dic, caption = "DIC comparison (observed-data log-likelihood)", digits = 2)
}
```

Interpretation: Lower DIC indicates better out-of-sample-leaning tradeoff
between fit and effective complexity within this model set. Large $\Delta$DIC
values imply the ranking is not a marginal tie.

DIC strongly prefers M3 over M2 ($\Delta$DIC = 2,220) and both over M1
($\Delta$DIC vs M1 $\approx$ 2.6 million). Effective parameters $p_D \approx 178.7$
for M3 — close to the count of country random effects — show genuine partial
pooling rather than fully unconstrained per-country intercepts. The DIC is
computed from the marginal beta-binomial log-likelihood (not JAGS's default
DIC, which conditions on latent $\theta_{it}$ for M2/M3 and is therefore not
directly comparable across models).

| $\Delta$DIC | Interpretation |
|-------------|----------------|
| > 10 | Strong evidence for lower-DIC model |
| 5–10 | Moderate evidence |
| < 5 | Interpret cautiously |


## How to Read the Model Comparison

Use this decision sequence:

1. **Start with PPC calibration:** If a model cannot reproduce key observed
  features (mean, variance, lower tail), do not treat it as adequate even if
  some coefficients look significant.
2. **Then check DIC gaps:** Use $\Delta$DIC to rank models among those that are
  at least approximately calibrated.
3. **Then read interpretability:** Prefer the model whose parameters map to
  meaningful data-generating components (baseline effects, overdispersion,
  persistent country heterogeneity).

In this report, that sequence points to M3: best overall tradeoff among the
compared models, lowest observed-data DIC, and interpretable overdispersion
plus country-level structure, despite residual PPC miscalibration in the lower
tail and within-region variance checks.

---


# Robustness Checks

## Frequentist Comparison

Selected coefficient alignments are shown below.

### Selected Bayesian vs frequentist coefficient comparison

```{r frequentist-table}
freq_path <- file.path(TABLES_DIR, "bayesian_vs_frequentist_comparison.csv")
if (file.exists(freq_path)) {
  freq_comp <- read.csv(freq_path)
  html_table(freq_comp, caption = "Bayesian vs. Frequentist coefficient comparison")
}
```

The full comparison CSV is available at `src/outputs/tables/bayesian_vs_frequentist_comparison.csv`.

M1 (binomial GLM via `glm`) and M2 (beta-binomial via `VGAM::vglm`) were fitted
successfully and their coefficients agree closely in sign and magnitude with the
Bayesian counterparts; interval widths are broadly comparable, with small
differences depending on parameter and model. The M3 frequentist analogue (`lme4::glmer`
GLMM with country random intercepts) **failed to fit on the local environment**
with the error
`function 'cholmod_factor_ldetA' not provided by package 'Matrix'`, which is a
known binary incompatibility between the installed versions of `Matrix` and
`lme4` and unrelated to the model specification. Because the frequentist
comparison is a bonus and the primary recommendation rests on the Bayesian
DIC, PPC, posterior-inference, and diagnostics analyses, this is documented as a
non-fatal limitation rather than remediated. Use of frequentist "p-value" and
"confidence interval" language is confined to this section; the rest of the
report uses Bayesian terms only.


## Sensitivity Analyses

```{r sensitivity-table}
sens_path <- file.path(TABLES_DIR, "sensitivity_summary.csv")
if (file.exists(sens_path)) {
  sens <- read.csv(sens_path, stringsAsFactors = FALSE)

  sens$sensitivity <- sub("^14\\.1", "S21.1", sens$sensitivity)
  sens$sensitivity <- sub("^14\\.2", "S21.2", sens$sensitivity)
  sens$sensitivity <- sub("^14\\.3", "S21.4", sens$sensitivity)
  sens$sensitivity <- sub("^14\\.4", "S21.5", sens$sensitivity)
  sens$sensitivity <- sub("^14\\.5", "S21.3", sens$sensitivity)

  sens_214 <- grepl("^S21\\.4", sens$sensitivity) & grepl("phi", sens$sensitivity, ignore.case = TRUE)
  sens_215 <- grepl("^S21\\.5", sens$sensitivity) & grepl("sigma", sens$sensitivity, ignore.case = TRUE)

  sens$status[sens_214] <-
   "SPECIFICATION ONLY - Bayesian refit not run"
  sens$status[sens_215] <-
   "SPECIFICATION ONLY - Bayesian refit not run"

  sens <- sens[order(sens$sensitivity), ]
  sens <- sens[, c("sensitivity", "status", "notes")]
  rownames(sens) <- NULL
  names(sens) <- c("Sensitivity arm", "Status", "Notes")
  html_table(sens, caption = "Summary of sensitivity analyses")
}
```

This sensitivity analysis includes five robustness arms with two execution levels.

1. **S21.1 Cohort threshold (executed):** `cohort > 0` vs. main `cohort >= 50`;
  frequentist M1 refit with stable coefficient direction/magnitude.
2. **S21.2 TB-HIV predictor (executed):** add standardized `e_tbhiv_prct`;
  frequentist M1 refit with a negative TB-HIV coefficient and broadly stable
  main effects on the reduced complete-case sample.
3. **S21.3 Post-2021 definitions (executed):** 2020–2023 subset using
  `used_2021_defs_flg == 1`; frequentist M1 refit with preserved main signs
  and wider uncertainty due reduced sample size.
4. **S21.4 Alternative $\phi$ prior (specification-only):** Bayesian prior arm
  prepared through alternative JAGS specification, but full Bayesian refit not
  included in the final computational run.
5. **S21.5 Alternative $\sigma_u$ prior (specification-only):** Bayesian prior
  arm prepared through alternative JAGS specification, but full Bayesian refit
  not included in the final computational run.

Three sensitivity checks were executed using frequentist analogues. Two
additional Bayesian prior-sensitivity arms were specified through alternative
JAGS model files, but full Bayesian refits were not executed in the final
computational run. Therefore, these prior checks are reported as
planned/specification-only rather than as completed Bayesian robustness
evidence.

The headline preference for M3 rests on DIC and PPC on the locked dataset and is
not contingent on the unexecuted Bayesian prior-sensitivity arms S21.4/S21.5.

---


# Discussion

## Discussion

### Research Question Answer

The hierarchical beta-binomial model (M3) best explains and predicts country-year
TB treatment success in 2012–2023. M1 fails: ordinary binomial sampling
variability cannot remotely account for the observed cross-country dispersion
($\Delta$DIC vs M3 $\approx$ 2.6 million; PPC variance test p $\approx$ 0). M2
improves substantially by adding extra-binomial overdispersion ($\phi$), but
$\Delta$DIC vs M3 is still 2,220 — strong evidence — and M2 still misfits the
cohort-weighted mean (p = 0.005) and over-predicts the lower tail. M3 is the
best overall model among the three candidates: it most successfully captures
central tendency (T1b posterior predictive p = 0.32) and approximately recovers
overall variance (T2 p = 0.04), while substantial residual structure is still
handled through country random intercepts. PPCs nevertheless show remaining
lower-tail and within-region variance miscalibration.

### Preferred Model Rationale

The technical case for M3 has three legs. First, **DIC** is decisive ($\Delta$DIC
> 10 is "strong evidence" by the standard convention; we observe $\Delta$DIC =
2,220 vs M2). Second, **posterior predictive checks** independently corroborate
the DIC ranking while still indicating residual lower-tail and within-region
variance miscalibration. Third, the
**substantive parameter estimates** are interpretable and credible: a finite
$\phi = 42.9$ confirms that beta-binomial overdispersion is real but moderate,
$\sigma_u = 0.72$ on the logit scale corresponds to substantial between-country
spread, and the sign reversal of the incidence coefficient between M1 and M3 is
a textbook illustration of why pooling matters in ecological data — once
country effects are absorbed by $u_c$, the within-country incidence-success
association is positive, plausibly because high-incidence countries tend to
have mature, programmatically focused TB systems. Remaining PPC miscalibration
in M3 is visible in both lower-tail frequency (T3 p = 1.0) and within-region
variance checks (low T4a/T4b p-values), making these natural targets for future
model extension.

### Limitations

1. **Ecological fallacy:** Country-level data cannot support individual-level
   causal claims.

2. **Reporting heterogeneity:** Countries differ in data quality, surveillance
   capacity, and conventions.

3. **Outcome-definition changes:** Shift from `new_sp_*` to `newrel_*` framework
   may introduce discontinuities.

4. **Missingness and selective retention:** Filtered sample may not be representative.

5. **Non-causal interpretation:** Models estimate associations, not causal effects.


## Conclusion

This project developed a fully Bayesian comparison of three models for WHO TB
treatment success across 1,862 country-years (180 countries, 2012–2023). The
hierarchical beta-binomial model (M3) is the recommended model. Binomial sampling
variability alone cannot explain the observed cross-country variation; adding a
beta-binomial overdispersion parameter ($\phi$) materially improves fit but
still misses the central tendency and the lower tail; only the further addition
of country-level random intercepts ($u_c$ with $\sigma_u \approx 0.72$ on the
logit scale) provides the best overall fit across DIC, central tendency,
variance, within-region heterogeneity, and country-level structure, although
residual lower-tail and within-region variance miscalibration remains. Mortality
remains negatively associated with success after pooling; incidence reverses to
a positive association once persistent country effects are absorbed; the
intercept implies a baseline success probability around 0.85. The full Bayesian
workflow — prior predictive checks, MCMC convergence diagnostics, observed-data
DIC, posterior predictive checks, and a 30/30/10-replicate parameter recovery
study — supports this recommendation. The frequentist comparison agrees on M1
and M2 sign and magnitude; the M3 GLMM analogue could not be fitted on the
local environment due to a Matrix/lme4 binary incompatibility, which we report
honestly as a non-fatal limitation of the bonus comparison.
Sensitivity evidence is supplementary: three executed frequentist checks and two
Bayesian prior-sensitivity specifications without final-run Bayesian refits;
the main conclusion does not depend on those unexecuted prior arms.

---


# Reproducibility

## Course Requirements Traceability

| Guideline requirement | Where addressed |
|---|---|
| Meaningful title | Title page |
| Dataset and goals | Sections 4-6 |
| Dataset description and at least two alternative models | Sections 5 and 8-11 |
| Role of parameters and inferential goals | Sections 8-12 and 16 |
| Parameter recovery with simulated data | Section 15 |
| MCMC output illustration | Sections 13-14 |
| Bayesian estimation / hypothesis testing / prediction | Sections 16-17 |
| Model comparison through DIC | Section 18 |
| Ability to recover observed-data features | Section 17 |
| Bonus: model checking diagnostics | Sections 14 and 17 |
| Bonus: frequentist comparison | Section 20 (partial: M1/M2 completed, M3 GLMM failed locally) |
| Sensitivity analyses (supplementary robustness) | Section 21 (partial: 3 frequentist checks executed; 2 Bayesian prior arms specification-only) |

Note: Partial status in Sections 20-21 reflects executed versus
specification-only supplementary work and does not change the primary Bayesian
model recommendation.


## Reproducibility Appendix

### Software Versions

```{r session-info}
sessionInfo()
```

### Data Provenance

- Source: World Health Organization Global TB Programme
- Files downloaded: 2026-04-04
- Analysis window: 2012-2023
- Final sample: 1,862 country-years, 180 countries


### File Structure

- Main analysis script: `src/main.R`
- JAGS model files: `src/models/*.jags`
- Output tables: `src/outputs/tables/`
- Output figures: `src/outputs/figures/`

### How to regenerate the final HTML report

The final HTML report is generated from:

src/report/report.Rmd

Command-line render:

<pre><code>Rscript -e "rmarkdown::render('src/report/report.Rmd', output_format = 'html_document', output_file = 'Arman_Feili_FSL2_Final_Report.html', output_dir = 'src/report')"</code></pre>
<!-- Rscript -e "rmarkdown::render('src/report/report.Rmd', output_format = 'html_document', output_file = 'Arman_Feili_FSL2_Final_Report.html', output_dir = 'src/report')" -->

R console render:

<pre><code>rmarkdown::render(
  "src/report/report.Rmd",
  output_format = "html_document",
  output_file = "Arman_Feili_FSL2_Final_Report.html",
  output_dir = "src/report"
)</code></pre>
<!-- output_file = "Arman_Feili_FSL2_Final_Report.html" -->

The full analysis source is src/main.R; JAGS model files are in src/models/.

### Final Validation Checks

These tables are reproducibility and submission-integrity checks, not
statistical evidence. Final validation verifies consistency of year window,
sample size, baseline region, predictor set, model numbering, and the
observed-data DIC method. The submission checklist marks the final HTML report
artifact as ready. The final written report artifact is
`src/report/Arman_Feili_FSL2_Final_Report.html`; any legacy duplicate HTML file is
not kept as a final artifact. The existing PDF export should be treated as
secondary unless regenerated successfully from the final Rmd source.

**Consistency, submission checklist, and version validation tables**

```{r final-validation-checks, results="asis"}
consistency_path <- file.path(TABLES_DIR, "consistency_check_results.csv")
checklist_path <- file.path(TABLES_DIR, "submission_package_checklist.csv")
version_path <- file.path(TABLES_DIR, "version_manifest.csv")

if (file.exists(consistency_path)) {
  consistency <- read.csv(consistency_path, stringsAsFactors = FALSE)
  html_table(consistency, caption = "Final consistency check results")
}

if (file.exists(checklist_path)) {
  checklist <- read.csv(checklist_path, stringsAsFactors = FALSE)

  target_row <- checklist$item == "Final written report (HTML)"
  if (any(target_row)) {
    final_html_rel <- "src/report/Arman_Feili_FSL2_Final_Report.html"
    final_html_abs <- file.path(PROJECT_ROOT, "report", "Arman_Feili_FSL2_Final_Report.html")
    checklist$location[target_row] <- final_html_rel
    checklist$exists[target_row] <- file.exists(final_html_abs)
    checklist$status[target_row] <- ifelse(checklist$exists[target_row], "Ready", "Missing")
  }

  html_table(checklist, caption = "Submission package checklist")
}

if (file.exists(version_path)) {
  version <- read.csv(version_path, stringsAsFactors = FALSE)
  html_table(version, caption = "Version manifest for final environment")
}
```


## References

