---
title: "Processed WHO TB Treatment Success — public analysis outline"
author: "Arman Feili"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# About this notebook

This is a **public summary** notebook for the Kaggle dataset
"Processed WHO TB Treatment Success Dataset, 2012-2023". It uses only
the processed, locked CSV that ships with the dataset and a few small
published tables. It is intentionally light: it does **not** rerun the
full Markov chain Monte Carlo (MCMC) workflow. The full Bayesian
beta-binomial fits (M1 pooled binomial, M2 beta-binomial, M3
hierarchical with country and region effects) were run in the original
project environment using R + JAGS, which is computationally expensive
and not always practical to set up inside a Kaggle notebook.

What you can do here:

- Load the locked analysis table and check its shape.
- Look at success-rate distributions, by-region and by-year patterns.
- Inspect summary tables of the published model comparison (DIC),
  posterior predictive checks (PPC), and the M3 posterior summary.

What you should **not** do here:

- Use this notebook as the source of truth for the published numbers —
  the locked CSV and the small bundled result tables are the source of truth.
- Treat the country-level success rates as performance rankings.

# Load packages and data

```{r packages}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
```

```{r load-data}
# Adjust the path if you have mounted the dataset under a different prefix.
data_path <- "../input/processed-who-tb-treatment-success-2012-2023"
if (!dir.exists(data_path)) data_path <- "."  # fallback for local runs

main_csv <- file.path(data_path, "processed_who_tb_treatment_success_country_year_2012_2023.csv")
df <- readr::read_csv(main_csv, show_col_types = FALSE)

cat("Rows:    ", nrow(df), "\n")
cat("Cols:    ", ncol(df), "\n")
cat("Years:   ", paste(range(df$year), collapse = " - "), "\n")
cat("Countries:", dplyr::n_distinct(df$iso3), "\n")
```

# Sample summary

```{r summary}
df %>%
  summarise(
    n_rows       = dplyr::n(),
    n_countries  = dplyr::n_distinct(iso3),
    n_years      = dplyr::n_distinct(year),
    median_cohort = median(cohort),
    median_success_rate = median(success / cohort)
  )
```

# EDA: success rate distribution

```{r eda-distribution}
df <- df %>% dplyr::mutate(prop = success / cohort)

ggplot(df, aes(x = prop)) +
  geom_histogram(bins = 40) +
  labs(
    title = "Distribution of treatment success rate (success / cohort)",
    x = "Success rate",
    y = "Country-years"
  ) +
  theme_minimal()
```

# EDA: success rate by WHO region

```{r eda-by-region}
ggplot(df, aes(x = g_whoregion, y = prop)) +
  geom_boxplot() +
  labs(
    title = "Treatment success rate by WHO region",
    x = "WHO region",
    y = "Success rate"
  ) +
  theme_minimal()
```

# EDA: trend by year

```{r eda-by-year}
df %>%
  dplyr::group_by(year) %>%
  dplyr::summarise(
    median_prop = median(prop),
    q25 = quantile(prop, 0.25),
    q75 = quantile(prop, 0.75),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = year, y = median_prop)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.2) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = seq(2012, 2023, 1)) +
  labs(
    title = "Median success rate per year (IQR shaded)",
    x = "Year",
    y = "Success rate"
  ) +
  theme_minimal()
```

# Published model results (read-only)

The full MCMC was run in the original project. The Kaggle dataset only
ships the processed analysis table and a few small documentation tables;
the published result tables (DIC, PPC, M3 posterior) are referenced here
for context. If they are not present in your working directory, the
chunks below will simply be skipped.

```{r dic, eval=file.exists('dic_comparison_table.csv')}
dic <- readr::read_csv("dic_comparison_table.csv", show_col_types = FALSE)
dic
```

```{r ppc, eval=file.exists('ppc_summary_table.csv')}
ppc <- readr::read_csv("ppc_summary_table.csv", show_col_types = FALSE)
ppc
```

```{r posterior-m3, eval=file.exists('posterior_summary_m3.csv')}
post_m3 <- readr::read_csv("posterior_summary_m3.csv", show_col_types = FALSE)
post_m3
```

# Notes on reproducibility

- This notebook only consumes the processed CSV; it does not refit
  models. To reproduce the published M1 / M2 / M3 fits you need R, the
  `rjags` package, and a JAGS installation.
- The locked CSV is sufficient to refit any of the three models, since
  it carries `success`, `cohort`, the standardized predictors, and the
  region / country indices.
- See `DATA_CARD.md`, `DATA_DICTIONARY.md`, and
  `PROVENANCE_AND_FILTERING.md` in the dataset folder for the full
  filtering and standardization story.

# Disclaimer

This is a processed derivative dataset, **not** an official WHO dataset
and **not** endorsed by WHO. Original source: WHO Global Tuberculosis
Programme. Country-level success rates should not be presented as
programme rankings.
