---
title: "Bayesian Modeling of WHO TB Treatment Success, 2012-2023"
subtitle: "A processed-derivative analysis built on the Kaggle public dataset"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: paged
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 110
)
```

```{r packages, include=FALSE}
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(forcats)
  library(scales)
  library(knitr)
})

find_file_by_name <- function(rel_or_name) {
  bn <- basename(rel_or_name)
  search_roots <- c(
    "/kaggle/input/datasets/armanfeili7/processed-who-tb-treatment-success-2012-2023",
    "/kaggle/input/processed-who-tb-treatment-success-2012-2023",
    "/kaggle/input",
    "../dataset",
    "../../Kaggle/dataset",
    "Kaggle/dataset",
    "."
  )
  for (root in search_roots) {
    if (dir.exists(root)) {
      cand <- file.path(root, rel_or_name)
      if (file.exists(cand)) return(cand)
    }
  }
  for (root in search_roots) {
    if (dir.exists(root)) {
      hits <- list.files(root, pattern = paste0("^", bn, "$"),
                         recursive = TRUE, full.names = TRUE)
      if (length(hits)) return(hits[[1]])
    }
  }
  NULL
}

read_csv_safe <- function(rel_or_name) {
  p <- find_file_by_name(rel_or_name)
  if (is.null(p)) return(NULL)
  readr::read_csv(p, show_col_types = FALSE)
}

show_table <- function(df, caption = NULL, digits = 3) {
  if (is.null(df)) return(invisible(NULL))
  knitr::kable(df, caption = caption, digits = digits)
}

base_theme <- function() {
  theme_minimal(base_size = 11) +
    theme(plot.title = element_text(face = "bold"),
          plot.subtitle = element_text(color = "grey30"))
}

model_palette <- c(M1 = "#C44E52", M2 = "#DD8452", M3 = "#55A868")
```

# Abstract

This notebook is the public companion to the Kaggle dataset
[`armanfeili7/processed-who-tb-treatment-success-2012-2023`](https://www.kaggle.com/datasets/armanfeili7/processed-who-tb-treatment-success-2012-2023).
It studies country-year TB treatment success in 1,862 country-years across 180
countries from 2012 to 2023, using a processed analysis table derived from
public WHO Global Tuberculosis Programme files. The original project compares
three Bayesian models — a binomial logistic regression, a beta-binomial model
with overdispersion, and a hierarchical beta-binomial model with country
random effects — using DIC and posterior predictive checks. The notebook
reproduces the public exploratory analysis and the compact model-comparison
layer using only the files shipped in the Kaggle dataset.

# Main finding

The hierarchical beta-binomial model (M3) is the preferred model. It has the
lowest DIC (24,940 vs 27,160 for M2 and ~2.67 million for M1) and the best
posterior predictive calibration among the three models. It captures both
overdispersion ($\phi \approx 42.9$) and persistent country-level
heterogeneity ($\sigma_u \approx 0.72$) that ordinary binomial sampling
variation alone cannot explain.

# Introduction

Tuberculosis remains a leading cause of death from a single infectious agent.
National TB programmes report treatment outcomes to the World Health
Organization each year, and treatment success is one of the headline
indicators of programme performance. WHO publishes descriptive statistics on
these outcomes, but country-year success rates vary in a structured way that
descriptive statistics alone cannot summarize: rates are concentrated at high
values, the lower tail is heavy, and regional patterns differ across years.

# Research question

> Which Bayesian model best explains and reproduces country-year TB treatment
> success from 2012 to 2023: a simple binomial logistic model, a beta-binomial
> model with overdispersion, or a hierarchical beta-binomial model with
> country random effects?

The question is answered using DIC for model comparison and posterior
predictive checks (PPC) on six test statistics. All models are fitted to the
same locked country-year sample so the comparison is fair.

> The full JAGS MCMC fitting was performed in the original project
> environment. This Kaggle notebook does not refit the models; instead, it
> uses the exported compact result tables included in the dataset to
> reproduce the public model-comparison and interpretation layer.

# Data and EDA

```{r load-main}
df <- read_csv_safe("processed_who_tb_treatment_success_country_year_2012_2023.csv")
stopifnot(!is.null(df))

stopifnot(
  nrow(df) == 1862L, ncol(df) == 17L,
  length(unique(df$iso3)) == 180L,
  min(df$year) == 2012L, max(df$year) == 2023L,
  all(df$success <= df$cohort),
  all(df$cohort >= 50),
  all(abs(df$prop_success - df$success / df$cohort) < 1e-9)
)
```

## Dataset inventory

```{r overview-table}
overall_cw_success <- sum(df$success) / sum(df$cohort)
show_table(
  tibble::tibble(
    Statistic = c(
      "Rows (country-years)", "Columns", "Distinct countries (iso3)",
      "Year range", "WHO regions",
      "Total successes", "Total cohort", "Cohort-weighted success rate"
    ),
    Value = c(
      format(nrow(df), big.mark = ","), as.character(ncol(df)),
      as.character(length(unique(df$iso3))),
      paste(min(df$year), max(df$year), sep = "-"),
      paste(sort(unique(df$g_whoregion)), collapse = ", "),
      format(sum(df$success), big.mark = ","),
      format(sum(df$cohort), big.mark = ","),
      sprintf("%.2f%%", 100 * overall_cw_success)
    )
  ),
  "Dataset overview"
)
```

The unit of analysis is a country-year. The locked sample contains nearly all
WHO regions across 12 consecutive years, with a global cohort-weighted
success rate close to 86%. The headline number is high, but the underlying
country-year distribution is far from uniform, which motivates the
model-comparison study.

```{r provenance}
prov <- read_csv_safe("documentation_tables/data_provenance.csv")
show_table(prov, "Raw WHO inputs used to build the processed table")
```

Two WHO files are the primary inputs: TB treatment outcomes (response) and
TB burden estimates (predictors). The data dictionary is used only as a
reference. The notification files are documented for completeness but they
are not part of the model inputs.

## Locked table schema

The locked table is the sole input for all three Bayesian models. The
response is the count of treatment successes out of the treatment cohort
(`success` and `cohort`). The proportion `prop_success` is included for
descriptive summaries and posterior predictive checks; the models are fitted
to counts. The continuous predictors used in the models are the standardized
versions (`year_z`, `e_inc_100k_z`, `e_mort_100k_z`, `c_cdr_z`). WHO region
is a categorical predictor with AFR as the reference baseline. The variables
`e_tbhiv_prct` and `used_2021_defs_flg` are kept for sensitivity analyses
only.

```{r variable-dict}
vd <- read_csv_safe("documentation_tables/project_variable_dictionary.csv")
show_table(vd, "Variable dictionary: roles and primary-vs-sensitivity usage")
```

```{r std-meta}
stdmeta <- read_csv_safe("documentation_tables/standardization_metadata.csv")
show_table(stdmeta, "Means and SDs used to build the standardized _z predictors")
```

## Filtering pipeline

```{r attrition}
attrition <- read_csv_safe("documentation_tables/attrition_table.csv")
show_table(attrition, "Sample attrition pipeline")
```

```{r attrition-plot, fig.height=4.2}
attrition %>%
  mutate(filter = fct_inorder(filter)) %>%
  ggplot(aes(x = fct_rev(filter), y = rows)) +
  geom_col(fill = "#4C72B0") +
  geom_text(aes(label = comma(rows)), hjust = -0.05, size = 3.4) +
  coord_flip() +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Sample attrition",
       subtitle = "Each row shows how many country-years remain after one filter",
       x = NULL, y = "Country-years remaining") +
  base_theme()
```

The pipeline reduces the merged raw rows from 5,130 to a locked sample of
1,862. The most substantial removals are non-comparable outcome definitions
(`rel_with_new_flg`), missing core predictors, and the cohort-size threshold
(`cohort >= 50`) that removes country-years where the empirical proportion
would be too noisy to model reliably.

## Final sample

```{r final-sample}
snapshot <- read_csv_safe("documentation_tables/final_sample_snapshot.csv")
show_table(snapshot, "Final sample snapshot")
```

The cohort-size filter removes 26 countries entirely. The most affected
region is AMR, where 19% of countries are lost. The same locked sample is
used for M1, M2, and M3, which is why the DIC comparison is fair.

## Exploratory patterns

```{r eda-prop-dist}
ggplot(df, aes(x = prop_success)) +
  geom_histogram(bins = 40, fill = "#4C72B0", color = "white") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "Distribution of treatment success rate",
       subtitle = "success / cohort, locked analysis sample (n = 1,862)",
       x = "Success rate", y = "Country-years") +
  base_theme()
```

The distribution is concentrated at high treatment success rates, but the
lower tail shows that some country-years have substantially poorer outcomes.
This motivates models that can handle more variability than a simple
binomial likelihood.

```{r eda-cohort, fig.height=3.8}
ggplot(df, aes(x = cohort)) +
  geom_histogram(bins = 50, fill = "#55A868", color = "white") +
  scale_x_log10(labels = comma) +
  labs(title = "Cohort size distribution (log10 x-axis)",
       x = "Cohort size", y = "Country-years") +
  base_theme()
```

Cohort sizes span several orders of magnitude. Modeling counts rather than
proportions preserves this information: country-years with very large
cohorts contribute much more to the likelihood than small ones.

```{r eda-region}
ggplot(df, aes(x = g_whoregion, y = prop_success, fill = g_whoregion)) +
  geom_boxplot(alpha = 0.85, outlier.alpha = 0.4) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(title = "Success rate by WHO region",
       x = "WHO region (AFR = reference)", y = "Success rate") +
  base_theme()
```

Regional distributions are not identical. AFR, EMR, SEA, and WPR cluster in
the mid-80% range, while EUR and AMR are noticeably lower. This supports
including WHO region as a categorical predictor, while still allowing
country-specific heterogeneity through the hierarchical model.

```{r eda-year-overall}
yearly <- df %>%
  group_by(year) %>%
  summarise(median_prop = median(prop_success),
            q25 = quantile(prop_success, 0.25),
            q75 = quantile(prop_success, 0.75),
            cohort_weighted = sum(success) / sum(cohort),
            .groups = "drop")

ggplot(yearly, aes(x = year)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.2, fill = "#4C72B0") +
  geom_line(aes(y = median_prop), color = "#4C72B0", linewidth = 1) +
  geom_point(aes(y = median_prop), color = "#4C72B0", size = 2) +
  geom_line(aes(y = cohort_weighted), color = "#C44E52", linewidth = 1, linetype = "dashed") +
  scale_x_continuous(breaks = seq(2012, 2023)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "Yearly success rate",
       subtitle = "Blue line + IQR band = country-year median; red dashed = cohort-weighted global rate",
       x = "Year", y = "Success rate") +
  base_theme()
```

The cohort-weighted global rate is consistently a few percentage points
higher than the country-year median. Both lines are broadly flat over the
study window, with only modest year-to-year movement.

```{r eda-region-trend}
trend_region <- df %>%
  group_by(g_whoregion, year) %>%
  summarise(mean_prop = sum(success) / sum(cohort), .groups = "drop")

ggplot(trend_region, aes(x = year, y = mean_prop, color = g_whoregion)) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.6) +
  scale_color_brewer(palette = "Set2", name = "WHO region") +
  scale_x_continuous(breaks = seq(2012, 2023)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "Temporal trend by WHO region",
       subtitle = "Cohort-weighted regional success rate per year",
       x = "Year", y = "Success rate") +
  base_theme()
```

```{r eda-heatmap, fig.height=3.6}
ggplot(trend_region, aes(x = year, y = g_whoregion, fill = mean_prop)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "#FDDBC7", high = "#1A9850",
                      labels = percent_format(accuracy = 1),
                      name = "Success rate") +
  scale_x_continuous(breaks = seq(2012, 2023)) +
  labs(title = "Region-year mean success rate",
       x = "Year", y = "WHO region") +
  base_theme()
```

Regional trends are not parallel. Some regions improve steadily, others
remain near a stable level, and a few show year-to-year noise. The heatmap
makes the persistent regional gaps visible at a glance.

```{r eda-country, fig.height=7}
country_summary <- df %>%
  group_by(iso3, g_whoregion) %>%
  summarise(mean_prop = sum(success) / sum(cohort), .groups = "drop")

top_bot <- bind_rows(
  country_summary %>% slice_max(mean_prop, n = 15) %>%
    mutate(panel = "Top 15 (highest mean success)"),
  country_summary %>% slice_min(mean_prop, n = 15) %>%
    mutate(panel = "Bottom 15 (lowest mean success)")
)

ggplot(top_bot, aes(x = mean_prop, y = fct_reorder(iso3, mean_prop), fill = g_whoregion)) +
  geom_col(alpha = 0.9) +
  facet_wrap(~ panel, scales = "free_y") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_brewer(palette = "Set2", name = "WHO region") +
  labs(title = "Top and bottom 15 countries by mean success rate (2012-2023)",
       x = "Mean success rate", y = NULL) +
  base_theme() +
  theme(legend.position = "bottom")
```

The country-level spread is wider than the regional spread, with the bottom
countries reaching well below 50%. This is exactly the kind of variation a
hierarchical model is designed to capture: country-specific levels that
remain stable across years.

```{r eda-bivariate, fig.height=4}
pretty_pred <- c(e_inc_100k = "Incidence per 100k",
                 e_mort_100k = "Mortality per 100k",
                 c_cdr = "Case detection ratio")
bivariate_long <- df %>%
  select(prop_success, g_whoregion, year, e_inc_100k, e_mort_100k, c_cdr) %>%
  pivot_longer(c(e_inc_100k, e_mort_100k, c_cdr),
               names_to = "predictor", values_to = "value") %>%
  mutate(predictor = factor(pretty_pred[predictor], levels = pretty_pred))

ggplot(bivariate_long, aes(x = value, y = prop_success)) +
  geom_point(aes(color = g_whoregion), alpha = 0.45, size = 1.1) +
  geom_smooth(method = "loess", se = FALSE, color = "black", linewidth = 0.7) +
  facet_wrap(~ predictor, scales = "free_x", nrow = 1) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_brewer(palette = "Set2", name = "WHO region") +
  labs(title = "Burden predictors vs treatment success",
       subtitle = "Each point is one country-year; black line is a loess trend",
       x = NULL, y = "Success rate") +
  base_theme() +
  theme(legend.position = "bottom")
```

The bivariate relationships are noisy, weakly non-linear, and have wide
between-country spread. Mortality has the clearest negative relationship,
incidence is weakly negative on the raw scale, and case detection has only a
mild positive slope. The substantial vertical spread is again a signal that
country-level random effects will matter.

```{r eda-corr, fig.height=4.5}
corr_vars <- intersect(
  c("year", "e_inc_100k", "e_mort_100k", "c_cdr", "e_tbhiv_prct"),
  names(df)
)
as.data.frame(as.table(cor(df[corr_vars], use = "pairwise.complete.obs"))) %>%
  rename(var1 = Var1, var2 = Var2, r = Freq) %>%
  ggplot(aes(x = var1, y = var2, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3.2) +
  scale_fill_gradient2(low = "#C44E52", mid = "white", high = "#4C72B0",
                       midpoint = 0, limits = c(-1, 1), name = "Pearson r") +
  labs(title = "Predictor correlation matrix",
       x = NULL, y = NULL) +
  base_theme() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Incidence and mortality are strongly correlated (r ≈ 0.84), but still below
the pre-specified exclusion threshold of 0.85. The remaining pairs are
weaker. The original project retained all four primary predictors together,
which is supported by this matrix.

```{r missingness}
miss_tbl <- df %>%
  summarise(across(everything(), ~ sum(is.na(.)) / n())) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_frac") %>%
  arrange(desc(missing_frac))
show_table(miss_tbl, "Missing-value fraction per column", digits = 4)
```

Primary predictors are complete by construction (the filtering pipeline
already dropped rows with missing core predictors). `used_2021_defs_flg` and
`e_tbhiv_prct` retain meaningful missingness, which is why they are kept for
sensitivity analyses only.

# Model roadmap

| Model | Likelihood | Added feature | Question answered |
|---|---|---|---|
| **M1** | Binomial | None | Is binomial sampling variance enough? |
| **M2** | Beta-Binomial | Overdispersion $\phi$ | Is there extra-binomial variation? |
| **M3** | Beta-Binomial + country REs | Country effects $u_i \sim N(0, \sigma_u^2)$ | Are there persistent country differences? |

## M1: Binomial

M1 is a logistic regression on the success counts with no extra variation.
It assumes that, after controlling for predictors and region, the only
remaining noise is ordinary binomial sampling variance. If treatment success
varies more than binomial sampling alone implies, M1 will systematically
underestimate variability.

## M2: Beta-Binomial

M2 keeps the same linear predictor as M1 but uses a beta-binomial likelihood
with a precision parameter $\phi$. Smaller $\phi$ corresponds to more
extra-binomial variation. As $\phi \to \infty$, M2 collapses to M1. M2 tests
whether allowing overdispersion improves fit.

## M3: Hierarchical beta-binomial

M3 keeps the beta-binomial likelihood and adds a country random effect
$u_i \sim N(0, \sigma_u^2)$ on top of the linear predictor. This separates
two sources of extra variation: $\phi$ for general overdispersion and
$\sigma_u$ for persistent country-level heterogeneity not explained by
predictors or region. M3 is the preferred model in the original analysis.

# Model comparison

```{r dic-table}
dic <- read_csv_safe("bayesian_result_tables/dic_comparison_table.csv")
show_table(dic, "DIC model comparison (lower is better)")
```

```{r dic-plot, fig.height=4}
dic_plot_data <- dic %>% mutate(Model = factor(Model, levels = c("M1", "M2", "M3")))

ggplot(dic_plot_data, aes(x = Model, y = DIC, fill = Model)) +
  geom_col(width = 0.6, alpha = 0.9) +
  geom_text(aes(label = comma(round(DIC))), vjust = -0.4, size = 3.6) +
  scale_y_continuous(labels = comma, trans = "log10",
                     expand = expansion(mult = c(0, 0.15))) +
  scale_fill_manual(values = model_palette, guide = "none") +
  labs(title = "DIC across models (log10 scale)",
       subtitle = "M1 is severely inadequate; M3 has the lowest DIC",
       x = NULL, y = "DIC") +
  base_theme()
```

```{r delta-dic, fig.height=4}
ggplot(dic_plot_data, aes(x = Model, y = Delta_DIC, fill = Model)) +
  geom_col(width = 0.6, alpha = 0.9) +
  geom_text(aes(label = comma(round(Delta_DIC))), vjust = -0.4, size = 3.6) +
  scale_fill_manual(values = model_palette, guide = "none") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Delta DIC relative to the preferred model (M3)",
       x = NULL, y = "Delta DIC") +
  base_theme()
```

M3 has the lowest DIC. The large difference between M3 and M2 indicates that
country-level random effects add important explanatory structure beyond
overdispersion alone, and the difference between M2 and M1 shows that
binomial sampling is far too rigid for this country-year panel.

# Posterior predictive checks

```{r ppc-table}
ppc <- read_csv_safe("bayesian_result_tables/ppc_summary_table.csv")
show_table(ppc, "Bayesian p-values per model and test statistic")
```

```{r ppc-plot, fig.height=5}
ggplot(ppc, aes(x = p_value, y = fct_rev(test_quantity), color = model)) +
  geom_vline(xintercept = 0.5, color = "grey50", linetype = "dashed") +
  geom_vline(xintercept = c(0.05, 0.95), color = "grey70", linetype = "dotted") +
  geom_point(size = 3, alpha = 0.85) +
  scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.05, 0.5, 0.95, 1)) +
  scale_color_manual(values = model_palette, name = "Model") +
  labs(title = "Posterior predictive Bayesian p-values",
       subtitle = "Values near 0.5 = well calibrated; values near 0 or 1 = systematic mismatch",
       x = "Bayesian p-value", y = "Test quantity") +
  base_theme() +
  theme(legend.position = "bottom")
```

```{r ppc-obs-rep, fig.height=4.5}
ggplot(ppc, aes(x = observed, y = rep_mean, color = model)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 2.5, alpha = 0.85) +
  scale_color_manual(values = model_palette, name = "Model") +
  labs(title = "Observed vs posterior predictive mean per PPC statistic",
       x = "Observed", y = "Posterior predictive mean") +
  base_theme() +
  theme(legend.position = "bottom")
```

M1 fails to reproduce the observed variance and lower tail. M2 improves
substantially, but M3 provides the best overall calibration among the three
models: the cohort-weighted mean is well calibrated ($p \approx 0.32$), the
variance is much closer to centered ($p \approx 0.04$), and the observed
versus posterior-predictive mean points sit closest to the 1:1 line. Some
lower-tail miscalibration remains under M3.

> Full posterior predictive draws and PPC histograms are not included in
> this compact Kaggle dataset, so density overlays from the original report
> are summarized here through the exported `ppc_summary_table.csv`.

# M3 posterior inference

```{r m3-posterior}
posterior_m3 <- read_csv_safe("bayesian_result_tables/posterior_summary_m3.csv")
show_table(posterior_m3, "M3 posterior summary (mean, sd, 95% credible interval)")
```

```{r m3-posterior-plot, fig.height=5.5}
posterior_m3 %>%
  mutate(label = fct_reorder(label, mean)) %>%
  ggplot(aes(x = mean, y = label)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_segment(aes(x = ci_lower, xend = ci_upper, y = label, yend = label),
               color = "#4C72B0", linewidth = 0.8) +
  geom_point(color = "#4C72B0", size = 2.4) +
  labs(title = "M3 posterior means with 95% credible intervals",
       subtitle = "Hierarchical beta-binomial model (logit scale)",
       x = "Posterior mean (CrI = 95% credible interval)", y = NULL) +
  base_theme()
```

Mortality has a clearly negative effect on treatment success (mean ≈ -0.39,
95% CrI excludes zero). Incidence has a positive effect once country effects
are absorbed (mean ≈ +0.28). The sign flip between the raw bivariate plots
and the M3 posterior is a between-country confounding signal: persistent
low-performing countries also tend to have high incidence, so the
conditional within-country association changes direction. Year and case
detection are near zero. The overdispersion parameter $\phi \approx 42.9$
confirms extra-binomial variance is present, and $\sigma_u \approx 0.72$
confirms substantial country-level heterogeneity beyond predictors and
region.

> The Kaggle dataset does not include the per-country random-effect table or
> the full posterior draws, so the country-level caterpillar plot from the
> original report is not reproduced here. The posterior summary above covers
> the M3 fixed effects, region effects, $\phi$, and $\sigma_u$ only.

# Interpretation

TB treatment success varies by country, region, and burden context.
Country-year data depart from a simple binomial model in two distinct ways:
extra-binomial dispersion within country-years and persistent differences
between countries. M2 addresses the first, M3 addresses both. M3 is
preferred not because of complexity but because both sources of variation
are real and measurable in the data.

# Limitations

- Observational country-level data; findings are associative, not causal.
- National TB programmes differ in reporting practices and data quality.
- WHO burden indicators (incidence, mortality, case detection) are
  estimates, not direct measurements.
- The raw WHO files are not redistributed in the Kaggle dataset; users who
  need them should download them from WHO.
- The full JAGS MCMC is not rerun here; the notebook uses only the exported
  compact result tables.
- Per-country random effects and full posterior predictive draws are not
  included in this compact dataset, so per-country rankings and PPC density
  overlays from the original report cannot be reproduced exactly.

# Reuse and citation

Please cite the World Health Organization Global Tuberculosis Programme as
the original data source, together with this processed Kaggle derivative
dataset. This is **not** an official WHO dataset and should not be treated
as an authoritative WHO publication.

# Conclusion

Three Bayesian models were fitted to the same locked country-year sample.
M1 (binomial) is severely inadequate, M2 (beta-binomial) closes most of the
gap by allowing overdispersion, and M3 (hierarchical beta-binomial) closes
the remainder by adding persistent country effects. **M3 is the preferred
model** in the original study, and the compact result tables shipped in the
Kaggle dataset reproduce that conclusion at the public level used in this
notebook.
